---
title: "Extended Data – NAO, JSL, and ENSO"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

Two-panel Nature-style figure:
- **Panel a** – ENSO vs NAO and ENSO vs JSL time series (stacked)
- **Panel b** – Impulse responses of log prices to ENSO, NAO, and JSL shocks
  (Fish prices and Grain prices × total effect vs. controlling for other indices)

## Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(fixest)
  library(purrr)
  library(glue)
  library(stringr)
})

source("../emileRegs.R")
setFixest_notes(FALSE)
```

## Data

```{r data}
fishprice <- read.csv("../processed data/fishprice_enso.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  dplyr::select(LocationSpecies, Year, Decade, everything()) %>%
  arrange(LocationSpecies, Year) %>%
  group_by(LocationSpecies) %>%
  mutate(logprice_diff = logprice - lag(logprice)) %>%
  ungroup()

price <- read.csv("../processed data/price_2023_enso.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  dplyr::select(Location, Year, Decade, everything()) %>%
  drop_na() %>%
  distinct() %>%
  arrange(Location, Year) %>%
  group_by(Location) %>%
  mutate(logprice_diff = logprice - lag(logprice)) %>%
  ungroup()
```

## Nature-style theme

```{r theme}
nature_theme <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      axis.line         = element_line(linewidth = 0.4, colour = "black"),
      axis.ticks        = element_line(linewidth = 0.35, colour = "black"),
      axis.ticks.length = unit(2.5, "pt"),
      axis.text         = element_text(size = base_size - 1, colour = "black"),
      axis.title        = element_text(size = base_size, colour = "black"),
      legend.position   = "bottom",
      legend.key.size   = unit(9, "pt"),
      legend.text       = element_text(size = base_size - 1),
      legend.title      = element_blank(),
      legend.margin     = margin(0, 0, 0, 0),
      legend.spacing.x  = unit(4, "pt"),
      strip.background  = element_blank(),
      strip.text        = element_text(face = "bold", size = base_size),
      plot.tag          = element_text(face = "bold", size = base_size + 2),
      plot.tag.position = "topleft",
      plot.margin       = margin(4, 6, 4, 6, "pt")
    )
}
```

## Panel a – Time series

```{r ts-data}
ts_data <- price %>%
  group_by(Year) %>%
  summarise(
    ENSO = mean(nino34,  na.rm = TRUE),
    NAO  = mean(NAO_cal, na.rm = TRUE),
    JSL  = mean(JSL,     na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(c(ENSO, NAO, JSL), ~ as.numeric(scale(.x))))

cor_nao <- cor.test(ts_data$ENSO, ts_data$NAO)
cor_jsl <- cor.test(ts_data$ENSO, ts_data$JSL)

fmt_p <- function(p) if (p < 0.001) "< 0.001" else sprintf("= %.3f", p)

lab_nao <- sprintf("r = %.2f, p %s", cor_nao$estimate, fmt_p(cor_nao$p.value))
lab_jsl <- sprintf("r = %.2f, p %s", cor_jsl$estimate, fmt_p(cor_jsl$p.value))

cat("ENSO vs NAO:", lab_nao, "\n")
cat("ENSO vs JSL:", lab_jsl, "\n")
```

```{r panel-a}
COL_ENSO <- "black"   # near-black
COL_NAO  <- "dodgerblue"   # vermillion
COL_JSL  <- "tomato"   # blue

df_nao <- ts_data %>%
  dplyr::select(Year, ENSO, NAO) %>%
  pivot_longer(-Year, names_to = "Index", values_to = "value") %>%
  mutate(Index = factor(Index, levels = c("ENSO", "NAO")))

df_jsl <- ts_data %>%
  dplyr::select(Year, ENSO, JSL) %>%
  pivot_longer(-Year, names_to = "Index", values_to = "value") %>%
  mutate(Index = factor(Index, levels = c("ENSO", "JSL")))

ylim_range <- range(c(df_nao$value, df_jsl$value), na.rm = TRUE)
ylim_pad   <- ylim_range + c(-0.08, 0.10) * diff(ylim_range)

make_ts_panel <- function(df, col_index, col_val, label_text,
                          x_title = NULL, y_title = "Standardized index") {
  ggplot(df, aes(Year, value, colour = Index)) +
    geom_hline(yintercept = 0, colour = "grey75", linewidth = 0.3, linetype = "dashed") +
    geom_line(linewidth = 0.55) +
    annotate(
      "text",
      x = min(ts_data$Year), y = ylim_pad[2],
      label = label_text,
      hjust = 0, vjust = 1, size = 4, colour = "grey35"
    ) +
    scale_colour_manual(values = col_index, labels = names(col_index)) +
    coord_cartesian(ylim = ylim_pad) +
    labs(x = x_title, y = y_title) +
    nature_theme() +
    guides(colour = guide_legend(override.aes = list(linewidth = 1.2)))
}

p_nao <- make_ts_panel(
  df_nao,
  col_index  = c(ENSO = COL_ENSO, NAO = COL_NAO),
  label_text = lab_nao
)

p_jsl <- make_ts_panel(
  df_jsl,
  col_index  = c(ENSO = COL_ENSO, JSL = COL_JSL),
  label_text = lab_jsl,
  x_title    = "Year"
)

panel_a <- wrap_plots(p_nao / p_jsl)
panel_a
```

## Panel b – Impulse responses

```{r irf-run, cache=TRUE}
COL_TOTAL <- "dodgerblue2"   # blue  – total effect
COL_CTRL  <- "tomato"   # red   – controlling for other indices

run_irfs <- function(data_in, var, controls_full, controls_base,
                     label, dataset_name, panel_id, fe) {
  bind_rows(
    lp_panel(
      data         = data_in,
      outcome      = "logprice",
      main_var     = var,
      controls     = controls_full,
      horizon      = 10,
      fe           = fe,
      panel_id     = panel_id,
      vcov_formula = DK ~ Year
    ) %>% mutate(shock = label, spec = "Controlling for others", dataset = dataset_name),

    lp_panel(
      data         = data_in,
      outcome      = "logprice",
      main_var     = var,
      controls     = controls_base,
      horizon      = 10,
      fe           = fe,
      panel_id     = panel_id,
      vcov_formula = DK ~ Year
    ) %>% mutate(shock = label, spec = "Total effect", dataset = dataset_name)
  )
}

shocks <- tribble(
  ~var,       ~label,  ~excl,
  "nino34",   "ENSO",  "NAO_cal + JSL",
  "NAO_cal",  "NAO",   "nino34 + JSL",
  "JSL",      "JSL",   "nino34 + NAO_cal"
) %>%
  mutate(
    # Add 3 lags of the shock to both specifications
    controls_full = glue("{excl} + l({var}, 1:3)"),
    controls_base = glue("l({var}, 1:3)")
  )

## Scale predictors to 1 SD so coefficients represent a 1-SD shock
sd_vars <- c("nino34", "NAO_cal", "JSL")

fishprice_sc <- fishprice %>%
  mutate(across(all_of(sd_vars), ~ .x / sd(.x, na.rm = TRUE)))

price_sc <- price %>%
  mutate(across(all_of(sd_vars), ~ .x / sd(.x, na.rm = TRUE)))

datasets <- list(
  list(data = fishprice_sc, name = "Fish prices",   panel_id = c("LocationSpecies", "Year"), fe = "LocationSpecies + Decade"),
  list(data = price_sc,     name = "Grain prices",  panel_id = c("Location", "Year"),        fe = "Location + Decade")
)

irf_all <- map_dfr(datasets, function(d) {
  pmap_dfr(shocks, function(var, label, excl, controls_full, controls_base) {
    run_irfs(d$data, var, controls_full, controls_base, label, d$name, d$panel_id, d$fe)
  })
}) %>%
  mutate(
    shock   = factor(shock,   levels = c("ENSO", "NAO", "JSL")),
    spec    = factor(spec,    levels = c("Total effect", "Controlling for others")),
    dataset = factor(dataset, levels = c("Fish prices", "Grain prices"))
  )
```

```{r panel-b, fig.width=5, fig.height=3.8}
panel_b <- ggplot(irf_all, aes(horizon, irf_mean, colour = spec, fill = spec, linetype = spec)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.13, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey55", linewidth = 0.3, linetype = "dashed") +
  geom_line(linewidth = 0.75) +
  facet_grid(dataset ~ shock) +
  scale_x_continuous(breaks = seq(0, 10, 2), expand = c(0.02, 0)) +
  scale_y_continuous(expand = c(0.07, 0)) +
  scale_colour_manual(values   = c(COL_TOTAL, COL_CTRL)) +
  scale_fill_manual(values     = c(COL_TOTAL, COL_CTRL)) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  labs(
    x = "Horizon (years)",
    y = "Effect on log prices"
  ) +
  guides(
    colour   = guide_legend(override.aes = list(linewidth = 1.2, alpha = 1)),
    fill     = guide_legend(override.aes = list(alpha = 0.2)),
    linetype = guide_legend(override.aes = list(linewidth = 1.2))
  ) +
  nature_theme()

panel_b
```

## Panel c – Famine onset probability by region

```{r famine-data}
famines <- read.csv("../processed data/famine_region_data.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  arrange(Region, Year) %>%
  mutate(across(all_of(sd_vars), ~ .x / sd(.x, na.rm = TRUE)))
```

```{r panel-c, fig.width=6, fig.height=3}
m_famine <- feols(
  Famine_start ~ NAO_cal:Region + JSL:Region + nino34:Region | Region + Decade,
  data     = famines,
  panel.id = c("Region", "Year"),
  vcov     = DK ~ Year
)

# Extract tidy coefficients for the three indices.
# fixest orders interaction terms inconsistently: NAO_cal:Region... but Region...:JSL
# and Region...:nino34 — so we match on presence of the keyword anywhere in the term.
coef_df <- broom::tidy(m_famine, conf.int = TRUE) %>%
  filter(grepl("nino34|NAO_cal|JSL", term)) %>%
  mutate(
    Index = case_when(
      grepl("nino34",  term) ~ "ENSO",
      grepl("NAO_cal", term) ~ "NAO",
      grepl("JSL",     term) ~ "JSL"
    ),
    # Strip index name and "Region" keyword to get the bare region string.
    # Handles both orderings: "NAO_cal:RegionX" and "RegionX:nino34"
    Region = term %>%
      gsub("NAO_cal|nino34|JSL", "", .) %>%   # remove index name
      gsub("Region",             "", .) %>%   # remove "Region" word
      gsub("^:|:$|::",          "", .) %>%   # remove leading/trailing colons
      trimws(),
    Index  = factor(Index, levels = c("ENSO", "NAO", "JSL"))
  )
region_order <- sort(unique(coef_df$Region))

coef_df <- coef_df %>%
  mutate(Region = factor(Region, levels = region_order))

panel_c <- ggplot(coef_df, aes(Region, estimate, colour = Index, shape = Index)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.35, linetype = "dashed") +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.18, linewidth = 0.5,
    position = position_dodge(width = 0.55)
  ) +
  geom_point(
    size = 1.8,
    position = position_dodge(width = 0.55)
  ) +
  scale_colour_manual(values = c(ENSO = COL_ENSO, NAO = COL_NAO, JSL = COL_JSL)) +
  scale_shape_manual(values  = c(ENSO = 16, NAO = 17, JSL = 15)) +
  labs(
    x = NULL,
    y = "Effect on Famine Prob. per SD"
  ) +
  nature_theme() +
  #make x axis text bigger
  
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 10),
    legend.position = "right"
  )

panel_c
```

## Combined figure

```{r combined, fig.width=10, fig.height=7}
fig <- (panel_a | panel_b) / panel_c +
  plot_layout(heights = c(1.6, 1)) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag = element_text(face = "bold", size = 14))

fig
```

## Save

```{r save, message=FALSE}
out_dir <- "../analysis/output/figures/extended data"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(
  file.path(out_dir, "figED_NAO_JSL_ENSO.pdf"),
  plot   = fig,
  width  = 9,
  height = 8,
  units  = "in",
  device = cairo_pdf
)

message("Saved → ", file.path(out_dir, "figED_NAO_JSL_ENSO.pdf"))
```
