---
title: "Trade and Market Integration – Extended Data"
output: html_notebook
editor_options:
 chunk_output_type: inline
---

Does ENSO raise wheat prices through trade disruption, or only through local
climate? We merge the Sound Toll trade dataset with ENSO and climate variables
and estimate the effect of ENSO on wheat prices, progressively controlling for
trade volume and local climate.

```{r setup}
suppressPackageStartupMessages(suppressWarnings({
  library(dplyr)
  library(haven)
  library(fixest)
  library(ggplot2)
}))
setFixest_notes(FALSE)
```

# Data

```{r load}
# Sound Toll dataset: port × year panel
# lnwheatindex     : log wheat price index at destination port
# lnpassage_nr     : log number of ship passages (trade volume)
# lntax_total_gold : log total tax paid in gold (trade value)
# growseasontemp_n5  : local growing-season temperature (5-yr smooth)
# ngrowseasontemp_n5 : negative growing-season temp (cold shocks)
# DestPortCorr_num : port city
# year

trade <- read_dta('../processed data/trade_sound_toll_dataset.dta') %>%
  filter(year >= 1500, year <= 1800) %>%
  mutate(port = as.character(DestPortCorr_num),
         Decade = floor(year / 10) * 10)

# Merge ENSO from famine panel (year-level)
famines <- read.csv('../processed data/famine_region_data.csv')

enso_yr <- famines %>%
  group_by(Year) %>%
  summarise(nino34      = mean(nino34,      na.rm = TRUE),
            PDSI_europe = mean(PDSI_europe, na.rm = TRUE),
            .groups = "drop")

trade <- trade %>%
  left_join(enso_yr, by = c("year" = "Year")) %>%
  filter(!is.na(nino34))

cat("Observations:", nrow(trade), "\n")
cat("Ports:", n_distinct(trade$port), "\n")
cat("Years:", min(trade$year), "-", max(trade$year), "\n")
```

# ENSO effect on wheat prices, controlling for trade

Columns add controls progressively:
1. ENSO only (port FE)
2. + local growing-season temperature
3. + European PDSI
4. + trade volume (log passages)
5. + trade value (log tax)
6. All controls + decade FE

```{r models}
m1 <- feols(lnwheatindex ~ l(nino34,0:5) | port + Decade,
            data = trade, cluster = ~port, panel.id = ~port + year)

m2 <- feols(lnwheatindex ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 | port,
            data = trade, cluster = ~port)

m3 <- feols(lnwheatindex ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 +
              PDSI_europe | port,
            data = trade, cluster = ~port)

m4 <- feols(lnwheatindex ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 +
              PDSI_europe + lnpassage_nr | port,
            data = trade, cluster = ~port)

m5 <- feols(lnwheatindex ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 +
              PDSI_europe + lntax_total_gold | port,
            data = trade, cluster = ~port)

m6 <- feols(lnwheatindex ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 +
              PDSI_europe + lnpassage_nr | port + Decade,
            data = trade, cluster = ~port)

dict <- c(
  nino34               = "Nino3.4",
  growseasontemp_n5    = "Grow. season temp.",
  ngrowseasontemp_n5   = "Neg. grow. season temp.",
  PDSI_europe          = "PDSI (Europe)",
  lnpassage_nr         = "Ln(trade passages)",
  lntax_total_gold     = "Ln(trade tax)"
)

etable(m1, m2, m3, m4, m5, m6,
       dict       = dict)
```

# ENSO effect on trade volume

Does ENSO reduce trade passages? (First stage of the trade-channel mechanism)

```{r enso-trade}
t1 <- feols(lnpassage_nr ~ nino34 | port,
            data = trade, cluster = ~port)

t2 <- feols(lnpassage_nr ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 | port,
            data = trade, cluster = ~port)

t3 <- feols(lnpassage_nr ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 +
              PDSI_europe | port,
            data = trade, cluster = ~port)

t4 <- feols(lnpassage_nr ~ nino34 + growseasontemp_n5 + ngrowseasontemp_n5 +
              PDSI_europe | port + Decade,
            data = trade, cluster = ~port)

etable(t1, t2, t3, t4,
       dict       = dict)
```

# Time series plot

```{r plot, fig.width=9, fig.height=4}
ts <- trade %>%
  group_by(year) %>%
  summarise(
    nino34       = mean(nino34,       na.rm = TRUE),
    lnwheatindex = mean(lnwheatindex, na.rm = TRUE),
    lnpassage_nr = mean(lnpassage_nr, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(c(nino34, lnwheatindex, lnpassage_nr), scale))

ggplot(ts, aes(x = year)) +
  geom_col(aes(y = nino34, fill = nino34 > 0), alpha = 0.45, width = 1) +
  geom_line(aes(y = lnwheatindex, color = "Wheat price"), linewidth = 1) +
  geom_line(aes(y = lnpassage_nr, color = "Trade passages"), linewidth = 1, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "cornflowerblue"), guide = "none") +
  scale_color_manual(values = c("Wheat price" = "black", "Trade passages" = "steelblue")) +
  labs(x = "Year", y = "Standardised value", color = NULL,
       title = "ENSO, Sound Toll trade passages, and wheat prices (1591–1800)",
       subtitle = "Bars = ENSO anomaly (red = El Niño); lines = standardised means across ports") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

ggsave("../figures/plots/ENSO_trade_wheat_ts.pdf", device = cairo_pdf, width = 9, height = 4)
```
