---
title: "ENSO Seasonal Timing – Reviewer Response"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

**Question (Reviewer):** Does using an annual NINO3.4 value conflate developing
and decaying ENSO events? Do results change if harvests are linked to the
preceding NDJF peak (decaying) vs the developing JJA phase?

**Answer strategy:** Construct three shock definitions from the paper's own
annual NINO3.4 index (already used in fig3). Since only an annual reconstruction
is available, seasonal variants are approximated as:

| Label | Definition | Interpretation |
|---|---|---|
| `nino34_annual` | annual mean[t] (paper's nino34) | Current baseline — identical to fig3 |
| `nino34_NDJF`   | (annual[t-1] + annual[t]) / 2 | Preceding winter peak (NDJF straddles calendar years) |
| `nino34_JJA`    | annual[t] | Growing-season phase (same year; = annual baseline) |

Note: because only an annual series is available, JJA and Annual are identical.
The meaningful comparison is Annual vs NDJF[t-1/t], which tests whether linking
harvests to the *preceding* winter peak (decaying El Niño) changes results.

If IRFs are similar, the annual measure is a robust integrated proxy. If NDJF
diverges, it indicates the preceding winter peak drives the teleconnection.

## Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(fixest)
  library(purrr)
  library(stringr)
})
source("../emileRegs.R")
setFixest_notes(FALSE)
```

## Merge with yield data

```{r yield-data}
# Load yield data (paper's nino34 already in this file)
yield_base <- read.csv("../processed data/yield_ljungqvist_2025.csv") %>%
  mutate(decade = floor(Year / 10) * 10) %>%
  filter(Grain %in% c("Wheat", "Rye")) %>%
  arrange(VarLocationGrain, Year)

# Build seasonal variants from the paper's own nino34 (not Cook reconstruction)
# so that the "Annual (current)" IRF exactly matches fig3.
# NDJF[t-1/t]: average of preceding and current year — best approximation when
# only an annual series is available (NDJF straddles two calendar years).
# JJA[t]: same-year annual value (growing-season developing phase).
nino_seasonal <- yield_base %>%
  distinct(Year, nino34) %>%
  arrange(Year) %>%
  mutate(
    nino34_annual = nino34,
    nino34_NDJF   = (lag(nino34, 1) + nino34) / 2,
    nino34_JJA    = nino34
  )

yield <- yield_base %>%
  left_join(nino_seasonal %>% dplyr::select(Year, nino34_annual, nino34_NDJF, nino34_JJA),
            by = "Year")

cat("Wheat/Rye obs:", nrow(yield), "\n")
cat("NAs in NDJF shock (from lag):", sum(is.na(yield$nino34_NDJF)), "\n")
```

## Check: seasonal variants vs paper nino34

```{r cross-check}
# Verify that nino34_annual == nino34 (should be 1.0) and inspect NDJF/JJA coverage
check <- yield %>%
  distinct(Year, nino34, nino34_annual, nino34_NDJF) %>%
  drop_na()

cat("Corr nino34 vs nino34_annual (should be 1):",
    round(cor(check$nino34, check$nino34_annual), 4), "\n")
cat("Corr nino34 vs nino34_NDJF:",
    round(cor(check$nino34, check$nino34_NDJF), 3), "\n")
cat("Years with valid NDJF:", sum(!is.na(yield %>% distinct(Year, nino34_NDJF) %>% pull(nino34_NDJF))), "\n")
```

## IRF under two shock definitions

```{r irf-seasonal, cache=TRUE}
# JJA == Annual when only an annual series is available, so we compare
# the two meaningful variants: current-year annual vs preceding-winter NDJF.
shocks <- c(
  "Annual (current)"  = "nino34_annual",
  "NDJF[t-1/t] peak" = "nino34_NDJF"
)

irf_seasonal <- map_dfr(names(shocks), function(label) {
  var <- shocks[[label]]
  lp_panel(
    data         = yield %>% filter(!is.na(.data[[var]])),
    outcome      = "logyield",
    main_var     = var,
    controls     = "i(decade)",
    horizon      = 3,
    fe           = "VarLocationGrain",
    panel_id     = c("VarLocationGrain", "Year"),
    vcov_formula = DK ~ Year
  ) %>%
    mutate(shock_label = label, term = var)
})
```

```{r plot-seasonal, fig.width=5, fig.height=3}
ggplot(irf_seasonal, aes(horizon, irf_mean,
                          colour = shock_label, fill = shock_label,
                          linetype = shock_label)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up),
              alpha = 0.12, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.35, linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_colour_manual(values = c(
    "Annual (current)"  = "#333333",
    "NDJF[t-1/t] peak" = "#0072B2"
  )) +
  scale_fill_manual(values = c(
    "Annual (current)"  = "#333333",
    "NDJF[t-1/t] peak" = "#0072B2"
  )) +
  scale_linetype_manual(values = c(
    "Annual (current)"  = "solid",
    "NDJF[t-1/t] peak" = "dashed"
  )) +
  labs(
    x        = "Horizon (years after shock)",
    y        = "% Harvest response",
    colour   = NULL, fill = NULL, linetype = NULL,
    title    = "Harvest IRF: annual vs NDJF ENSO",
    subtitle = "Wheat & Rye | DK SEs | VarLocationGrain + Decade FE"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position  = "bottom",
    strip.background = element_blank()
  ) +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.2, alpha = 1)))
```

## Two-panel figure: teleconnected W/R vs non-teleconnected non-W/R

```{r irf-two-panels, cache=TRUE}
# Panel a: teleconnected Wheat & Rye (yield already filtered to W/R above)
# Panel b: non-teleconnected, non-Wheat/Rye — load full yield for other grains
yield_all <- read.csv("../processed data/yield_ljungqvist_2025.csv") %>%
  mutate(decade = floor(Year / 10) * 10) %>%
  arrange(VarLocationGrain, Year) %>%
  left_join(nino_seasonal %>% dplyr::select(Year, nino34_annual, nino34_NDJF),
            by = "Year")

run_irf <- function(df) {
  map_dfr(names(shocks), function(slabel) {
    var <- shocks[[slabel]]
    lp_panel(
      data         = df %>% filter(!is.na(.data[[var]])),
      outcome      = "logyield",
      main_var     = var,
      controls     = "i(decade)",
      horizon      = 3,
      fe           = "VarLocationGrain",
      panel_id     = c("VarLocationGrain", "Year"),
      vcov_formula = DK ~ Year
    ) %>% mutate(shock_label = factor(slabel, levels = names(shocks)))
  })
}

irf_teleco_WR   <- run_irf(yield %>% filter(teleco_PDSI_10 == 1))
irf_nonteleco_noWR <- run_irf(yield_all %>%
                                 filter(!Grain %in% c("Wheat", "Rye"),
                                        teleco_PDSI_10 == 0))
```

```{r plot-two-panels, fig.width=6.5, fig.height=2.8}
COL_ANN  <- "#E69F00"   # warm amber — distinct, colourful
COL_NDJF <- "#0072B2"   # blue

make_panel <- function(df, title) {
  ggplot(df, aes(horizon, irf_mean, colour = shock_label, fill = shock_label,
                 linetype = shock_label)) +
    geom_ribbon(
      data = df %>% filter(shock_label == "Annual (current)"),
      aes(ymin = irf_down, ymax = irf_up),
      alpha = 0.15, colour = NA
    ) +
    geom_hline(yintercept = 0, colour = "grey65", linewidth = 0.3, linetype = "dashed") +
    geom_line(linewidth = 0.75) +
    geom_point(size = 1.5) +
    scale_x_continuous(breaks = 0:3, expand = c(0.04, 0)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                       expand = c(0.06, 0)) +
    scale_colour_manual(
      values = c("Annual (current)"  = COL_ANN,
                 "NDJF[t-1/t] peak" = COL_NDJF),
      labels = c("Annual (current)"  = "Annual",
                 "NDJF[t-1/t] peak" = "NDJF peak")
    ) +
    scale_fill_manual(
      values = c("Annual (current)"  = COL_ANN,
                 "NDJF[t-1/t] peak" = COL_NDJF),
      labels = c("Annual (current)"  = "Annual",
                 "NDJF[t-1/t] peak" = "NDJF peak")
    ) +
    scale_linetype_manual(
      values = c("Annual (current)"  = "solid",
                 "NDJF[t-1/t] peak" = "dashed"),
      labels = c("Annual (current)"  = "Annual",
                 "NDJF[t-1/t] peak" = "NDJF peak")
    ) +
    labs(x = "Horizon (years)", y = "Harvest response (%)",
         colour = NULL, fill = NULL, linetype = NULL,
         title = title) +
    theme_classic(base_size = 8) +
    theme(
      legend.position   = "bottom",
      legend.key.width  = unit(18, "pt"),
      legend.text       = element_text(size = 7.5),
      legend.margin     = margin(0, 0, 0, 0),
      legend.spacing.x  = unit(4, "pt"),
      strip.background  = element_blank(),
      plot.title        = element_text(size = 8, face = "plain", colour = "grey30"),
      axis.line         = element_line(linewidth = 0.35),
      axis.ticks        = element_line(linewidth = 0.3),
      axis.ticks.length = unit(2.5, "pt"),
      plot.margin       = margin(4, 8, 4, 4, "pt")
    ) +
    guides(colour   = guide_legend(override.aes = list(linewidth = 1.1, alpha = 1)),
           fill     = guide_legend(override.aes = list(alpha = 0.2)),
           linetype = guide_legend(override.aes = list(linewidth = 1.1)))
}

# Shared y limits across both panels
combined_range <- range(c(irf_teleco_WR$irf_down,   irf_teleco_WR$irf_up,
                           irf_nonteleco_noWR$irf_down, irf_nonteleco_noWR$irf_up),
                        na.rm = TRUE)
y_pad  <- diff(combined_range) * 0.06
y_lims <- combined_range + c(-y_pad, y_pad)

pa <- make_panel(irf_teleco_WR,      "Teleconnected, Wheat & Rye")    + coord_cartesian(ylim = y_lims)
pb <- make_panel(irf_nonteleco_noWR, "Not teleconnected, other grains") + coord_cartesian(ylim = y_lims)

fig_save <- pa + pb +
  plot_layout(guides = "collect") &
  plot_annotation(tag_levels = "a") &
  theme(
    legend.position  = "bottom",
    plot.tag         = element_text(face = "bold", size = 9)
  )

fig_save
```

## Numerical comparison table

```{r table}
irf_seasonal %>%
  dplyr::select(shock_label, horizon, irf_mean, irf_down, irf_up) %>%
  mutate(across(c(irf_mean, irf_down, irf_up), ~ round(.x * 100, 2))) %>%
  pivot_wider(names_from = shock_label,
              values_from = c(irf_mean, irf_down, irf_up)) %>%
  arrange(horizon) %>%
  print()
```

## Save

```{r save, message=FALSE}
out_dir <- "../analysis/output/figures/extended data"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(
  file.path(out_dir, "figED_seasonal_ENSO_timing.pdf"),
  plot = fig_save, width = 6.5, height = 3.0,
  units = "in", device = cairo_pdf
)
message("Saved → figED_seasonal_ENSO_timing.pdf")
```

