---
title: "R Notebook"
output: html_notebook
editor_options:
 chunk_output_type: inline
---


```{r}
suppressPackageStartupMessages(suppressWarnings(library(ggh4x)))
suppressPackageStartupMessages(suppressWarnings(library(dplyr)))
suppressPackageStartupMessages(suppressWarnings(library(tidyr)))
suppressPackageStartupMessages(suppressWarnings(library(tidyverse)))
suppressPackageStartupMessages(suppressWarnings(library(ggplot2)))
suppressPackageStartupMessages(suppressWarnings(library(magrittr)))
suppressPackageStartupMessages(suppressWarnings(library(future)))
suppressPackageStartupMessages(suppressWarnings(library(stringr)))
suppressPackageStartupMessages(suppressWarnings(library(glue)))
suppressPackageStartupMessages(suppressWarnings(library(ggrepel)))     
suppressPackageStartupMessages(suppressWarnings(library(ggdist) ))     
suppressPackageStartupMessages(suppressWarnings(library(scales)))      
suppressPackageStartupMessages(suppressWarnings(library(patchwork)))
suppressPackageStartupMessages(suppressWarnings(library(maps)))
suppressPackageStartupMessages(suppressWarnings(library(sf)))
suppressPackageStartupMessages(suppressWarnings(library(gridExtra)))
suppressPackageStartupMessages(suppressWarnings(library(viridis)))
suppressPackageStartupMessages(suppressWarnings(library(fixest)))
suppressPackageStartupMessages(suppressWarnings(library(ivreg)))
suppressPackageStartupMessages(suppressWarnings(library(rms)))
suppressPackageStartupMessages(suppressWarnings(library(ggpubr)))
suppressPackageStartupMessages(suppressWarnings(library(ltm)))
suppressPackageStartupMessages(suppressWarnings(library(pglm)))
suppressPackageStartupMessages(suppressWarnings(library(data.table)))
suppressPackageStartupMessages(suppressWarnings(library(stargazer)))
suppressPackageStartupMessages(suppressWarnings(library(finalfit)))
suppressPackageStartupMessages(suppressWarnings(library(modelsummary)))
suppressPackageStartupMessages(suppressWarnings(library(ggtext)))
suppressPackageStartupMessages(suppressWarnings(library(useful)))
library(furrr)
library(purrr)
source("../emileRegs.R")
setFixest_notes(FALSE)

```

# Data
```{r}


famines <- read.csv("../processed data/famine_region_data.csv") %>% mutate(Decade = floor(Year/10)*10) %>% 
  #sort years within region
  arrange(Region, Year) 


fishprice <- read.csv('../processed data/fishprice_enso.csv') %>% 
  mutate(Decade = floor(Year / 10) * 10) %>%
  dplyr::select(LocationSpecies,Year, Decade, everything()) %>% 
  #create a logprice_diff variable
  arrange(LocationSpecies, Year) %>%
  group_by(LocationSpecies) %>%
  mutate(logprice_diff = logprice - dplyr::lag(logprice)) %>% 
  ungroup() 

price <- read.csv('../processed data/price_2023_enso.csv')
#put Location and Year as first columns
price <- price %>%
    mutate(Decade = floor(Year / 10) * 10) %>%
    dplyr::select(Location,Year,Decade, everything()) %>%
  drop_na() %>%
  distinct() %>%
  arrange(Location, Year) %>%
  group_by(Location) %>%
  mutate(logprice_diff = logprice - dplyr::lag(logprice)) %>%
  ungroup()

yield_2023 <- read.csv('../processed data/yield_ljungqvist_2025.csv') %>%
  mutate(decade = floor(Year / 10) * 10) %>%
  dplyr::select(VarLocationGrain, Year, decade, everything()) %>%
  mutate(dy = logyield - lag(logyield))

# v2 yield panel with raw logyield (cubic record trend in the FE clause
# replaces the per-record SD normalisation; β reads as % harvest).
yield_v2_GT3 <- read.csv('../processed data/yield_ljungqvist_v2.csv') %>%
  mutate(decade = floor(Year / 10) * 10,
         Year2  = Year^2,
         Year3  = Year^3) %>%
  filter(!is.na(yield), yield > 0)


```




# Non Linearity


```{r, fig.width=10, fig.height=4}

# Linear
m_lin <- feols(
  f(logprice, 1) - l(logprice, 1) ~ 
    nino34 + 
    i(Decade) | Location,
  data = price,
  panel.id = c("Location", "Year"),
  vcov = DK ~ Year
)

# Quadratic
m_p2 <- feols(
  f(logprice, 1) - l(logprice, 1) ~  poly(nino34,2,raw=T) + 
    i(Decade) | Location,
  data = price,
  panel.id = c("Location", "Year"),
  vcov = DK ~ Year
)

# Cubic
m_p3 <- feols(
  f(logprice, 1) - l(logprice, 1) ~ poly(nino34,3,raw=T) +
    i(Decade) | Location,
  data = price,
  panel.id = c("Location", "Year"),
  vcov = DK ~ Year
)

# Helper to generate predictions over a grid
make_pred <- function(model, label) {
  useful::predict_poly(
    model, "nino34",
    min = -1, max = 1.5,
    ref = 0, step.length = 0.01,
    xvar_name = "nino34"
  ) %>%
    mutate(spec = label)
}

# Predictions
rf_lin <- make_pred(m_lin, "Linear")
rf_p2  <- make_pred(m_p2,  "Quadratic")
rf_p3  <- make_pred(m_p3,  "Cubic")

# Combine
rf_all <- bind_rows(rf_lin, rf_p2, rf_p3) %>%
  mutate(spec = factor(spec, levels = c("Linear", "Quadratic", "Cubic")))

# Plot
ggplot(rf_all, aes(x = nino34, y = response)) +
  geom_line(color = "cornflowerblue", linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "cornflowerblue", alpha = 0.25) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~spec, nrow = 1) +
  labs(
    x = "NINO3.4 Index",
    y = "Estimated Effect on Price Growth"
  ) +
  theme_classic(base_size = 15)
ggsave("../figures/plots/ED_nonlin_price.pdf")
```





```{r, fig.width=10, fig.height=4}
#--- Estimate models ---
m_lin <- feols(
  Famine_start ~ nino34  | Region,
  data = famines %>% filter(Region == "Central Europe"),
  vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

m_p2 <- feols(
  Famine_start ~ poly(nino34, 2, raw = TRUE) |Region ,
  data = famines %>% filter(Region == "Central Europe"),
  vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

m_p3 <- feols(
  Famine_start ~ poly(nino34, 3, raw = TRUE) | Region,
  data = famines %>% filter(Region == "Central Europe"),
  vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

#--- Predictions ---
rf_lin <- useful::predict_poly(m_lin, "nino34", min = -1, max = 1.5,
                              ref = 0, step.length = 0.01,
                              xvar_name = "nino34") %>%
  mutate(spec = "Linear")

rf_p2 <- useful::predict_poly(m_p2, "nino34", min = -1, max = 1.5,
                             ref = 0, step.length = 0.01,
                             xvar_name = "nino34") %>%
  mutate(spec = "Quadratic")

rf_p3 <- useful::predict_poly(m_p3, "nino34", min = -1, max = 1.5,
                             ref = 0, step.length = 0.01,
                             xvar_name = "nino34") %>%
  mutate(spec = "Cubic")

rf_all <- bind_rows(rf_lin, rf_p2, rf_p3)%>%
  mutate(spec = factor(spec, levels = c("Linear", "Quadratic", "Cubic")))

#--- Plot ---
ggplot(rf_all, aes(x = nino34, y = response)) +
  geom_line(color = "cornflowerblue", size = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "cornflowerblue", alpha = 0.25) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~spec, nrow = 1) +
  labs(
    x = "NINO3.4 Index",
    y = "Estimated Effect on Famine Start Probability"
  ) +
  theme_classic(base_size = 15)
ggsave("../figures/plots/nonlinearity_onset.pdf", device = cairo_pdf)
```


## Non-linearity: 4 outcomes, linear (+ CI) vs quadratic vs cubic (estimates only)

```{r nonlin-4panel, fig.width=8, fig.height=6}

# ── predict_poly wrapper ────────────────────────────────────────────────────
pp <- function(m) useful::predict_poly(m, "nino34",
                                       min = -1, max = 1.5, ref = 0,
                                       step.length = 0.01, xvar_name = "nino34")

# ── Generic panel builder ───────────────────────────────────────────────────
# CI ribbon for linear only; quadratic/cubic: line only
# title uses plotmath for small-caps label prefix via expression()
make_panel <- function(lin, p2, p3, y_lab, tag) {
  spec_col <- c("Linear" = "#4477AA", "Quadratic" = "tomato", "Cubic" = "orange")

  df_lin <- lin %>% mutate(spec = "Linear")
  df_p2  <- p2  %>% mutate(spec = "Quadratic")
  df_p3  <- p3  %>% mutate(spec = "Cubic")

  ggplot() +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_ribbon(data = df_lin,
                aes(x = nino34, ymin = lower, ymax = upper),
                fill = spec_col["Linear"], alpha = 0.2) +
    geom_line(data = bind_rows(df_lin, df_p2, df_p3),
              aes(x = nino34, y = response, colour = spec), linewidth = 0.9) +
    scale_colour_manual(values = spec_col, name = NULL,
                        guide = guide_legend(override.aes = list(linewidth = 1.5))) +
    labs(x = "NINO3.4", y = y_lab, tag = tag) +
    theme_classic(base_size = 13) +
    theme(
      legend.position  = "none",
      plot.tag         = element_text(size = 18, face = "bold"),
      axis.title.y     = element_text(size = 10),
      panel.grid.minor = element_blank()
    )
}

# ══════════════════════════════════════════════════════════════════════════════
# Panel A: Famine onset – Central Europe
# ══════════════════════════════════════════════════════════════════════════════
ce <- famines %>% filter(Region == "Central Europe")

m1_lin <- feols(Famine_start ~ nino34                  | Region,
                data = ce, panel.id = c("Region","Year"), vcov = DK ~ Year)
m1_p2  <- feols(Famine_start ~ poly(nino34,2,raw=TRUE)  | Region,
                data = ce, panel.id = c("Region","Year"), vcov = DK ~ Year)
m1_p3  <- feols(Famine_start ~ poly(nino34,3,raw=TRUE)  | Region,
                data = ce, panel.id = c("Region","Year"), vcov = DK ~ Year)

p1 <- make_panel(pp(m1_lin), pp(m1_p2), pp(m1_p3),
                 y_lab = "P(famine onset)",
                 tag   = "a")

# ══════════════════════════════════════════════════════════════════════════════
# Panel B: Grain prices at h = 1
# ══════════════════════════════════════════════════════════════════════════════
m2_lin <- feols(f(logprice,1) - l(logprice,1) ~ nino34                  + i(Decade) | Location,
                data = price, panel.id = c("Location","Year"), vcov = DK ~ Year)
m2_p2  <- feols(f(logprice,1) - l(logprice,1) ~ poly(nino34,2,raw=TRUE) + i(Decade) | Location,
                data = price, panel.id = c("Location","Year"), vcov = DK ~ Year)
m2_p3  <- feols(f(logprice,1) - l(logprice,1) ~ poly(nino34,3,raw=TRUE) + i(Decade) | Location,
                data = price, panel.id = c("Location","Year"), vcov = DK ~ Year)

p2 <- make_panel(pp(m2_lin), pp(m2_p2), pp(m2_p3),
                 y_lab = "Log grain price change (h = 1)",
                 tag   = "b")

# ══════════════════════════════════════════════════════════════════════════════
# Panel C: WR teleconnected yields at h = 0 (v2 panel, gold-ticket #3 spec)
# Outcome: log(yield/sd(yield)) — β reads as % harvest (sd absorbed by VLG FE).
# FE: VarLocationGrain[Year, Year2, Year3] (cubic record trend).
# Controls: i(decade) + l(nino34, 1:3).
# ══════════════════════════════════════════════════════════════════════════════
WR_tele_nl <- yield_v2_GT3 %>% filter(Grain %in% c("Wheat","Rye"),
                                       teleco_PDSI_10 == 1)

m3_lin <- feols(f(logyield,0) - l(logyield,1) ~ nino34 + l(nino34, 1:3)
                + i(decade) | VarLocationGrain[Year, Year2, Year3],
                data = WR_tele_nl, panel.id = c("VarLocationGrain","Year"),
                vcov = DK ~ Year)
m3_p2  <- feols(f(logyield,0) - l(logyield,1) ~ poly(nino34,2,raw=TRUE)
                + l(nino34, 1:3) + i(decade)
                | VarLocationGrain[Year, Year2, Year3],
                data = WR_tele_nl, panel.id = c("VarLocationGrain","Year"),
                vcov = DK ~ Year)
m3_p3  <- feols(f(logyield,0) - l(logyield,1) ~ poly(nino34,3,raw=TRUE)
                + l(nino34, 1:3) + i(decade)
                | VarLocationGrain[Year, Year2, Year3],
                data = WR_tele_nl, panel.id = c("VarLocationGrain","Year"),
                vcov = DK ~ Year)

p3 <- make_panel(pp(m3_lin), pp(m3_p2), pp(m3_p3),
                 y_lab = "Log yield change (h = 0)",
                 tag   = "c")

# ══════════════════════════════════════════════════════════════════════════════
# Panel D: Fish prices at h = 3
# ══════════════════════════════════════════════════════════════════════════════
m4_lin <- feols(f(logprice,3) - l(logprice,1) ~ nino34                  + i(Decade) | LocationSpecies,
                data = fishprice, panel.id = c("LocationSpecies","Year"), vcov = DK ~ Year)
m4_p2  <- feols(f(logprice,3) - l(logprice,1) ~ poly(nino34,2,raw=TRUE) + i(Decade) | LocationSpecies,
                data = fishprice, panel.id = c("LocationSpecies","Year"), vcov = DK ~ Year)
m4_p3  <- feols(f(logprice,3) - l(logprice,1) ~ poly(nino34,3,raw=TRUE) + i(Decade) | LocationSpecies,
                data = fishprice, panel.id = c("LocationSpecies","Year"), vcov = DK ~ Year)

p4 <- make_panel(pp(m4_lin), pp(m4_p2), pp(m4_p3),
                 y_lab = "Log fish price change (h = 3)",
                 tag   = "d")

# ══════════════════════════════════════════════════════════════════════════════
# Shared legend (extracted from p1 with legend turned on temporarily)
# ══════════════════════════════════════════════════════════════════════════════
p_legend <- p1 +
  theme(
    legend.position     = "top",
    legend.text         = element_text(size = 14),
    legend.key.width    = unit(1.5, "cm"),
    legend.spacing.x    = unit(0.4, "cm")
  )

legend_grob <- ggpubr::get_legend(p_legend)

# ══════════════════════════════════════════════════════════════════════════════
# 2×2 layout with legend above
# ══════════════════════════════════════════════════════════════════════════════
p_grid <- (p1 | p2) / (p3 | p4)

p_all <- ggpubr::ggarrange(
  ggpubr::as_ggplot(legend_grob),
  p_grid,
  ncol    = 1,
  heights = c(0.06, 1)
)

p_all
ggsave("../analysis/output/figures/appendix/FigA_nonlin.pdf", p_all, device = cairo_pdf)
```


# TYW

```{r}
m_onset <- feols(
  Famine_start ~ nino34:Region + i(Decade)  | Region,
  data = famines,
  vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

m_tyw <- feols(
  Famine_start ~ nino34:Region + i(Decade)  | Region,
  data = famines %>% 
  filter(!between(Year, 1618, 1648)),
  vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

etable(m_onset, m_tyw, vcov = DK ~ Year,
       keep_raw = "nino34:RegionCentral",
       #change model names
       headers = c("Full sample", "Excl. Thirty Years' War"),
       tex = TRUE, replace= TRUE,
       file = "../analysis/output/tables/TYW_CE_nino34_coef.tex")
```

```{r}
etable(m_onset, m_tyw, keep_raw = "nino34:RegionCentral",
       vcov = DK ~ Year)
```

# Fig 2A as contingency table: Fisher's exact test

NINO3.4 dichotomised at 0 (El Niño vs neutral/La Niña) × Famine onset (yes/no),
Central Europe only.

```{r fisher-fig2a}
ce <- famines %>%
  filter(Region == "Central Europe") %>%
  mutate(
    ElNino      = ifelse(nino34 > 0, "NINO3.4 > 0", "NINO3.4 <= 0"),
    FamineOnset = ifelse(Famine_start == 1, "Famine", "No Famine")
  )

ct <- table(ElNino = ce$ElNino, FamineOnset = ce$FamineOnset)
print(ct)

# Reorder so NINO3.4 > 0 is the first row → OR > 1 means El Niño increases famine odds
ct_ord <- ct[c("NINO3.4 > 0", "NINO3.4 <= 0"), c("Famine", "No Famine")]

ft <- fisher.test(ct_ord, alternative = "greater")
cat(sprintf("\nFisher's exact test (one-sided: El Niño → more famines)\n"))
cat(sprintf("  Odds ratio: %.3f\n", ft$estimate))
cat(sprintf("  95%% CI:    [%.3f, Inf]\n", ft$conf.int[1]))
cat(sprintf("  p-value:   %.4f\n", ft$p.value))
```

```{r fisher-latex}
out_tab <- "../analysis/output/tables"
dir.create(out_tab, recursive = TRUE, showWarnings = FALSE)

lines <- c(
  "\\begin{table}[h]",
  "\\centering",
  paste0("\\caption*{Contingency table: NINO3.4 sign $\\times$ famine onset (Central Europe, 1500--1800). ",
         "Fisher's exact test (one-sided, El Ni\\~{n}o $\\to$ more famines): ",
         sprintf("odds ratio = %.2f (95\\%% CI: %.2f, $\\infty$), $p = %.4f$.}", ft$estimate, ft$conf.int[1], ft$p.value)),
  "\\label{tab:contingency_fisher}",
  "\\begin{tabular}{lcc}",
  "\\toprule",
  " & \\textbf{Famine onset} & \\textbf{No famine} \\\\",
  "\\midrule",
  sprintf("\\textbf{NINO3.4 $> 0$}    & %d & %d \\\\",
          ct_ord["NINO3.4 > 0",  "Famine"], ct_ord["NINO3.4 > 0",  "No Famine"]),
  sprintf("\\textbf{NINO3.4 $\\leq 0$} & %d & %d \\\\",
          ct_ord["NINO3.4 <= 0", "Famine"], ct_ord["NINO3.4 <= 0", "No Famine"]),
  "\\bottomrule",
  "\\end{tabular}",
  "\\end{table}"
)

writeLines(lines, file.path(out_tab, "contingency.tex"))
cat("Saved contingency.tex\n")
```


# Trade Market Integration (Sound Toll)

```{r trade-sound-toll}
# ── Load and aggregate Sound Toll data to yearly level ─────────────────────
sound_raw <- haven::read_dta("../processed data/trade_sound_toll_dataset.dta")

# Average log passages and log tax across ports within each year
sound_yearly <- data.frame(
  Year              = as.integer(names(tapply(sound_raw$lnpassage_nr,    sound_raw$year, mean, na.rm = TRUE))),
  ln_passages_total = as.numeric(tapply(sound_raw$lnpassage_nr,    sound_raw$year, mean, na.rm = TRUE)),
  ln_tax_total      = as.numeric(tapply(sound_raw$lntax_total_gold, sound_raw$year, mean, na.rm = TRUE))
)

# ── Merge with famine panel ─────────────────────────────────────────────────
famines_trade <- famines %>%
  left_join(sound_yearly, by = "Year")

cat("Famine obs with Sound Toll coverage:",
    sum(!is.na(famines_trade$ln_passages_total)), "/", nrow(famines_trade), "\n")
cat("Year range with Sound Toll data:",
    range(famines_trade$Year[!is.na(famines_trade$ln_passages_total)]), "\n")
```

```{r trade-lpm-models}
# ── Baseline: full sample ────────────────────────────────────────────────────
m_base_full <- feols(
  Famine_start ~ nino34:Region + i(Decade) | Region,
  data = famines, vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

# ── Restrict to Sound Toll overlap (1591–1800) ───────────────────────────────
famines_sound <- famines_trade %>% filter(!is.na(ln_passages_total))

# Baseline: restricted sample
m_base_restricted <- feols(
  Famine_start ~ nino34:Region + i(Decade) | Region,
  data = famines_sound, vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

# Base + trade (ln_passages + ln_tax)
m_trade <- feols(
  Famine_start ~ nino34:Region + i(Decade) +
    ln_passages_total + ln_tax_total | Region,
  data = famines_sound, vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

# Base + trade + climate/conflict controls
m_trade_ctrl <- feols(
  Famine_start ~ nino34:Region + i(Decade) +
    ln_passages_total + ln_tax_total +
    JSL:Region + NAO_cal:Region +
    ongoing_wars:Region + log(1 + Deaths):Region | Region,
  data = famines_sound, vcov = DK ~ Year,
  panel.id = c("Region", "Year")
)

etable(m_base_full, m_base_restricted, m_trade, m_trade_ctrl,
       keep_raw = "nino34:RegionCentral Europe")
```

```{r trade-table, results='asis'}
# ── Regression table ────────────────────────────────────────────────────────
etable(
  m_base_full, m_base_restricted, m_trade, m_trade_ctrl,
  keep_raw     = c("nino34:RegionCentral Europe"),
  drop.section = "fixef",
  fitstat      = c("N", "cor2"),
  dict         = c(
    "nino34:RegionCentral Europe" = "Nino3.4 $\\times$ Central Europe",
    ln_passages_total             = "ln(Sound Toll passages)",
    ln_tax_total                  = "ln(Sound Toll tax)",
    Famine_start                  = "Famine Start"
  ),
  extralines = list(
    "Decade FE"                  = c("Yes", "Yes", "Yes", "Yes"),
    "Region FE"                  = c("Yes", "Yes", "Yes", "Yes"),
    "Climate/Conflict Controls"  = c("No",  "No",  "No",  "Yes"),
    "Sound Toll Control"         = c("No",  "No",  "Yes", "Yes"),
    "Sample"                     = c("1500--1800", rep("1591--1800", 3))
  ),
  tex    = TRUE,
  label  = "tab:trade_onset",
  file   = "../analysis/output/tables/trade_onset.tex",
  title  = "Famine Onset Model with Trade Controls"
)
```


# Trade Market Integration – IRF Analysis

## Panel A: Grain Price IRF (baseline vs. controlling for trade & tax)

```{r price-trade-irf-data}
# Merge price panel with Sound Toll yearly data
price_trade <- price %>%
  left_join(sound_yearly, by = "Year") %>%
  filter(!is.na(ln_passages_total), !is.na(logprice))


```

```{r price-trade-irf-run}
# Baseline price IRF (restricted to Sound Toll overlap years)
irf_price_base <- lp_panel(
  data         = price,
  outcome      = "logprice",
  main_var     = "nino34",
  controls     = "i(Decade) + l(nino34, 1:3)",
  horizon      = 10,
  fe           = "Location",
  panel_id     = c("Location", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(label = "Baseline")

# Price IRF controlling for trade (ln_passages) and tax (ln_tax)
irf_price_trade <- lp_panel(
  data         = price_trade,
  outcome      = "logprice",
  main_var     = "nino34",
  controls     = "i(Decade) + ln_passages_total + ln_tax_total + l(nino34, 1:3)",
  horizon      = 10,
  fe           = "Location",
  panel_id     = c("Location", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(label = "Trade & Tax Controls")

irf_price_combined <- bind_rows(irf_price_base, irf_price_trade) %>%
  mutate(label = factor(label, levels = c("Baseline", "Trade & Tax Controls")))
```

```{r price-trade-irf-figure, fig.width=7, fig.height=5}
col_price <- c("Baseline" = "#4477AA", "Trade & Tax Controls" = "firebrick")

pA <- ggplot(irf_price_combined, aes(x = horizon, y = irf_mean,
                                      colour = label, fill = label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up),
              alpha = 0.10, colour = NA) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous("Horizon (years)", breaks = 0:10) +
  scale_y_continuous("log grain price variation") +
  scale_colour_manual(values = col_price, name = NULL) +
  scale_fill_manual(values   = col_price, name = NULL) +
  labs(tag = "A") +
  theme_classic(base_size = 15) +
  theme(
    plot.tag           = element_text(size = 16, face = "bold"),
    legend.position    = "bottom"
  )

pA
```


## Panel B: Grain Yield IRF (teleconnected vs. non-teleconnected, controlling for price)

```{r yield-trade-irf-data}


WR_ctrl <- yield_2023 %>%
  filter(Grain %in% c("Wheat", "Rye")) %>%
  left_join(sound_yearly, by = "Year")

WR_base <- yield_2023 %>%
  filter(Grain %in% c("Wheat", "Rye"))
```

```{r yield-trade-irf-run}
ctrl_yield_base  <- "i(decade) + l(nino34, 1:3)"
ctrl_yield_price <- "i(decade) + ln_passages_total + ln_tax_total + l(nino34, 1:3)"

WR_tele_ctrl <- WR_ctrl %>% filter(teleco_PDSI_10 == 1)
WR_weak_ctrl <- WR_ctrl %>% filter(teleco_PDSI_10 == 0)

WR_tele_base <- WR_base %>% filter(teleco_PDSI_10 == 1)
WR_weak_base <- WR_base %>% filter(teleco_PDSI_10 ==0)

irf_yield_tele_base <- lp_panel(
  data         = WR_tele_base %>% filter(!is.na(logyield)),
  outcome      = "logyield",
  main_var     = "nino34",
  controls     = ctrl_yield_base,
  horizon      = 5,
  fe           = "VarLocationGrain",
  panel_id     = c("VarLocationGrain", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(label = "Teleconnected")

irf_yield_weak_base <- lp_panel(
  data         = WR_weak_base,
  outcome      = "logyield",
  main_var     = "nino34",
  controls     = ctrl_yield_base,
  horizon      = 5,
  fe           = "VarLocationGrain",
  panel_id     = c("VarLocationGrain", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(label = "Non-Teleconnected")

irf_yield_tele_price <- lp_panel(
  data         = WR_tele_ctrl,
  outcome      = "logyield",
  main_var     = "nino34",
  controls     = ctrl_yield_price,
  horizon      = 5,
  fe           = "VarLocationGrain",
  panel_id     = c("VarLocationGrain", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(label = "Teleco + Price Control")

irf_yield_weak_price <- lp_panel(
  data         = WR_weak_ctrl,
  outcome      = "logyield",
  main_var     = "nino34",
  controls     = ctrl_yield_price,
  horizon      = 5,
  fe           = "VarLocationGrain",
  panel_id     = c("VarLocationGrain", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(label = "Non-Teleco + Price Control")

irf_yield_combined <- bind_rows(
  irf_yield_tele_base, irf_yield_weak_base,
  irf_yield_tele_price, irf_yield_weak_price
) %>%
  mutate(label = factor(label, levels = c(
    "Teleconnected", "Non-Teleconnected",
    "Teleco + Price Control", "Non-Teleco + Price Control"
  )))
```

```{r yield-trade-irf-figure, fig.width=7, fig.height=5}
col_yield <- c(
  "Teleconnected"              = "firebrick",
  "Non-Teleconnected"          = "cornflowerblue",
  "Teleco + Price Control"     = "#CC4444",
  "Non-Teleco + Price Control" = "#6688CC"
)
lty_yield <- c(
  "Teleconnected"              = "solid",
  "Non-Teleconnected"          = "solid",
  "Teleco + Price Control"     = "dashed",
  "Non-Teleco + Price Control" = "dashed"
)

offsets_y <- c(
  "Teleconnected"              = -0.15,
  "Non-Teleconnected"          = -0.05,
  "Teleco + Price Control"     =  0.05,
  "Non-Teleco + Price Control" =  0.15
)

irf_yield_plot <- irf_yield_combined %>%
  mutate(h = horizon + offsets_y[as.character(label)])

pB <- ggplot(irf_yield_plot, aes(x = h, y = irf_mean,
                                   colour = label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(ymin = irf_down, ymax = irf_up), linewidth = 0.8) +
  geom_point(aes(shape = label), size = 2.5) +
  scale_shape_manual(values = c(
    "Teleconnected"              = 16,
    "Non-Teleconnected"          = 16,
    "Teleco + Price Control"     = 17,
    "Non-Teleco + Price Control" = 17
  ), name = NULL,
     labels = c("Teleconnected", "Non-Teleconnected",
                "Teleco +\nTrade/Tax Control", "Non-Teleco +\nTrade/Tax Control")) +
  scale_x_continuous("Horizon (years)", breaks = 0:5) +
  scale_y_continuous("log grain harvest") +
  scale_colour_manual(values = col_yield, name = NULL,
                      labels = c("Teleconnected", "Non-Teleconnected",
                                 "Teleco +\nTrade/Tax Control", "Non-Teleco +\nTrade/Tax Control")) +
  labs(tag = "B") +
  theme_classic(base_size = 15) +
  theme(
    plot.tag        = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    legend.key.size = unit(0.4, "cm"),
    legend.text     = element_text(size = 11),
    legend.spacing.x = unit(0.3, "cm")
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE),
         shape  = guide_legend(nrow = 2, byrow = TRUE))

pB
```


## Combined two-panel figure

```{r trade-irf-combined-figure, fig.width=10, fig.height=5}
p_trade_irf <- pA + pB + plot_layout(ncol = 2)
p_trade_irf

ggsave("../analysis/output/figures/extended data/ED_trade_irf.pdf",
       p_trade_irf, device = cairo_pdf,
       create.dir = TRUE)
```


# Yield IRF – Wheat/Rye, Teleconnected regions

```{r yield-irf-setup}
source("../emileRegs.R")

WR <- yield_2023 %>% filter(Grain %in% c("Wheat", "Rye"))

# ── helpers (mirrors 03_fig3_harvest.R) ────────────────────────────────────
run_lp_wr <- function(df, endog, controls = "i(decade)", hor = 3) {
  lp_panel(
    data         = df,
    outcome      = endog,
    main_var     = "nino34",
    controls     = controls,
    horizon      = hor,
    fe           = "VarLocationGrain",
    panel_id     = c("VarLocationGrain", "Year"),
    vcov_formula = DK ~ Year
  )
}

plot_irf <- function(df_irf, title) {
  ggplot(df_irf, aes(x = horizon, y = irf_mean, ymin = irf_down, ymax = irf_up,
                     colour = label, fill = label)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_ribbon(alpha = 0.15, colour = NA) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    scale_x_continuous(breaks = 0:3) +
    labs(x = "Horizon (years)", y = "% change in log yield",
         title = title, colour = NULL, fill = NULL) +
    theme_classic(base_size = 13) +
    theme(legend.position = "bottom")
}
```

```{r yield-irf-run}
WR_all   <- WR
WR_weak  <- WR %>% filter(teleco_PDSI_10 == 0)
WR_tele  <- WR %>% filter(teleco_PDSI_10 == 1)

ctrl_base <- "i(decade)"
ctrl_full <- "i(decade) + JSL + NAO_cal + ongoing_wars + Deaths"
# The LP LHS is f(y,h) - l(y,1), so adding l(logyield,1) as a control
# is equivalent to including a lag of the dependent variable level
ctrl_lag  <- "i(decade) + l(dy, 1)"
ctrl_lag_full <- "i(decade) + l(dy, 1) + JSL + NAO_cal + ongoing_wars + Deaths"

# ── Baseline (replicates Fig 3C) ────────────────────────────────────────────
irf_base <- bind_rows(
  run_lp_wr(WR_all,  "logyield", controls = ctrl_base) %>% mutate(group = "All regions"),
  run_lp_wr(WR_weak, "logyield", controls = ctrl_base) %>% mutate(group = "Weakly affected"),
  run_lp_wr(WR_tele, "logyield", controls = ctrl_base) %>% mutate(group = "Teleconnected"),
  run_lp_wr(WR_tele, "logyield", controls = ctrl_full) %>% mutate(group = "Teleco + controls")
) %>% mutate(spec = "Baseline")

# ── Lag-augmented: adds l(logyield,1) to the RHS ───────────────────────────
irf_lag <- bind_rows(
  run_lp_wr(WR_all,  "logyield", controls = ctrl_lag)      %>% mutate(group = "All regions"),
  run_lp_wr(WR_weak, "logyield", controls = ctrl_lag)      %>% mutate(group = "Weakly affected"),
  run_lp_wr(WR_tele, "logyield", controls = ctrl_lag)      %>% mutate(group = "Teleconnected"),
  run_lp_wr(WR_tele, "logyield", controls = ctrl_lag_full) %>% mutate(group = "Teleco + controls")
) %>% mutate(spec = "Lag-augmented")

irf_all <- bind_rows(irf_base, irf_lag) %>%
  mutate(
    group = factor(group, levels = c("All regions", "Weakly affected",
                                     "Teleconnected", "Teleco + controls")),
    spec  = factor(spec,  levels = c("Baseline", "Lag-augmented"))
  )
```

```{r yield-irf-figure, fig.width=8, fig.height=4}
col_map <- c("All regions"       = "black",
             "Weakly affected"   = "cornflowerblue",
             "Teleconnected"     = "firebrick",
             "Teleco + controls" = "orange")

# small horizontal offset per group so errorbars don't overlap
offsets <- c("All regions" = -0.18, "Weakly affected" = -0.06,
             "Teleconnected" = 0.06, "Teleco + controls" = 0.18)

irf_plot <- irf_all %>%
  mutate(h = horizon + offsets[as.character(group)])

p <- ggplot(irf_plot, aes(x = h, y = irf_mean, colour = group)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(ymin = irf_down, ymax = irf_up), linewidth = 1.0) +
  geom_point(size = 2.8) +
  facet_wrap(~spec, nrow = 1) +
  scale_x_continuous("Horizon (years)", breaks = 0:3) +
  scale_y_continuous("% harvest response") +
  scale_colour_manual(values = col_map, name = NULL) +
  coord_cartesian(ylim = c(-0.16, 0.08)) +
  theme_classic(base_size = 14) +
  theme(
    strip.text         = element_text(size = 13, face = "bold"),
    strip.background   = element_blank(),
    legend.position    = "bottom",
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4)
  )

p
ggsave("../figures/plots/ED_yield_irf_lagcheck.pdf", p, device = cairo_pdf)
```


# Price IRF – Inflation rate as dependent variable

The main paper (Fig 5A) estimates the IRF of ENSO on **log price levels**
(`logprice`). Here we use the **log price change** (`logprice_diff` = year-on-year
inflation rate) as the outcome variable instead, so the IRF shows the effect on the
*rate* of price growth rather than the cumulative level.

```{r price-irf-inflation}
# LP variant without subtracting l(y,1): LHS is f(outcome, h) rather than
# f(outcome, h) - l(outcome, 1). Appropriate here because the outcome is
# already a first difference (logprice_diff), so no further differencing needed.
# Scale nino34 by its cross-sample SD so IRF is per 1 SD shock
nino34_sd  <- sd(price$nino34, na.rm = TRUE)

price_diff <- price %>%
  filter(!is.na(logprice_diff)) %>%
  mutate(nino34_scaled = nino34 / nino34_sd)

irf_grain_infl <- lp_panel(
  data         = price_diff,
  outcome      = "logprice_diff",
  main_var     = "nino34_scaled",
  controls     = "i(Decade)",
  horizon      = 10,
  fe           = "Location",
  panel_id     = c("Location", "Year"),
  vcov_formula = DK ~ Year
)
```

```{r price-irf-inflation-figure, fig.width=7, fig.height=5}
ggplot(irf_grain_infl, aes(x = horizon, y = irf_mean, ymin = irf_down, ymax = irf_up)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(fill = "cornflowerblue", alpha = 0.2) +
  geom_line(colour = "cornflowerblue", linewidth = 0.9) +
  geom_point(colour = "cornflowerblue", size = 2) +
  scale_x_continuous("Horizon (years)", breaks = seq(0, 10, 2)) +
  scale_y_continuous("Effect on log price inflation (per 1 SD ENSO shock)") +
  ggtitle("Grain price inflation response to ENSO") +
  theme_classic(base_size = 13) +
  theme(panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4))

ggsave("../figures/plots/ED_price_irf_inflation.pdf",
       width = 7, height = 5, device = cairo_pdf)
```

