---
title: "lp_panel vs lp_lin_panel – reconciliation"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

**Root cause:** 11 LocationSpecies series have year gaps (up to 73 years).
`dplyr::lead(y, h)` in `lp_lin_panel` shifts rows positionally, so a gap year
counts as h steps forward. `fixest`'s `f(y, h)` uses actual year values via
`panel.id`, so it correctly skips to calendar year t+h.

**Two fixes tested:**

1. **Drop gapped series** — run both estimators on the balanced subsample
2. **Expand to balanced panel** — fill gaps with `NA` rows so `dplyr::lead`
   lands on the same observations as `fixest`'s `f()`

Fix 2 is preferable: it uses all data in `lp_panel` while making `lp_lin_panel`
match exactly.

## Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(fixest)
  library(lpirfs)
  library(patchwork)
})
source("../emileRegs.R")
setFixest_notes(FALSE)
```

## Data + gap diagnosis

```{r data}
fishprice_raw <- read.csv("../processed data/fishprice_enso.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  arrange(LocationSpecies, Year)

# Identify series with any year gaps
gap_summary <- fishprice_raw %>%
  group_by(LocationSpecies) %>%
  mutate(gap = Year - dplyr::lag(Year)) %>%
  summarise(
    max_gap  = max(gap, na.rm = TRUE),
    n_gaps   = sum(gap > 1, na.rm = TRUE),
    n_obs    = n(),
    .groups  = "drop"
  ) %>%
  arrange(desc(max_gap))

cat("Series with gaps:\n")
print(gap_summary %>% filter(n_gaps > 0))

# Series with a gap >= 10 years are dropped: the within-entity mean is
# meaningless across such a break, and at horizon h fixest's f() would
# also cross the gap for some t. Threshold = max horizon (10 years).
keep_series <- gap_summary %>% filter(max_gap < 20) %>% pull(LocationSpecies)
drop_series <- gap_summary %>% filter(max_gap >= 20) %>% pull(LocationSpecies)

cat("\nDropped (max_gap >= 15 yrs):", length(drop_series), "\n")
print(drop_series)
cat("Kept:", length(keep_series), "\n")
```

## Fix – drop max_gap >= 10 series, then balance remaining gaps with NA rows

```{r fix-data}
# Step 1: drop series with large breaks
fishprice_dropped <- fishprice_raw %>%
  filter(LocationSpecies %in% keep_series)

# Step 2: fill remaining small gaps (< 10 yrs) with NA rows so dplyr::lead
# lands on the same calendar year as fixest's f()
fishprice_fixed <- fishprice_dropped %>%
  group_by(LocationSpecies) %>%
  complete(Year = full_seq(Year, 1)) %>%
  ungroup() %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  arrange(LocationSpecies, Year)

cat("Rows after fix:",  nrow(fishprice_fixed),
    "from", n_distinct(fishprice_fixed$LocationSpecies), "units\n")
cat("(vs", nrow(fishprice_raw), "original rows from",
    n_distinct(fishprice_raw$LocationSpecies), "units)\n")

cat("Rows after balancing:", nrow(fishprice_fixed),
    "(vs", nrow(fishprice_raw), "original)\n")
```

## Run all specs

```{r run, cache=TRUE, message=FALSE, warning=FALSE}
HOR <- 10

run_fixest <- function(df, label) {
  lp_panel(
    data         = df,
    outcome      = "logprice",
    main_var     = "nino34",
    controls     = NULL,
    horizon      = HOR,
    fe           = "LocationSpecies + Decade",
    panel_id     = c("LocationSpecies", "Year"),
    vcov_formula = DK ~ Year
  ) %>% mutate(spec = label)
}

# Matches plm exactly on full sample: pre-compute LHS using positional
# dplyr::lead/lag (grouped by entity, sorted by year), then regress with
# feols using no panel.id — so fixest treats f()/l() the same way plm does.
run_fixest_positional <- function(df, label) {
  df <- df %>% arrange(LocationSpecies, Year)

  map_dfr(0:HOR, function(h) {
    df_h <- df %>%
      group_by(LocationSpecies) %>%
      mutate(.lhs = dplyr::lead(logprice, h) - dplyr::lag(logprice, 1)) %>%
      ungroup() %>%
      filter(!is.na(.lhs), !is.na(nino34))

    fml <- as.formula(".lhs ~ nino34 | LocationSpecies + Decade")
    mod  <- feols(fml, data = df_h, vcov = DK ~ Year)
    beta <- coef(mod)["nino34"]
    se   <- sqrt(vcov(mod)["nino34", "nino34"])
    tibble(horizon = h, irf_mean = beta, se = se,
           irf_down = beta - 1.96*se, irf_up = beta + 1.96*se,
           spec = label)
  })
}

run_plm <- function(df, label) {
  res <- lp_lin_panel(
    data_set     = df %>%
      mutate(Decade = as.factor(Decade)) %>%
      dplyr::select(LocationSpecies, Year, logprice, nino34, Decade),
    endog_data   = "logprice",
    cumul_mult   = TRUE,
    shock        = "nino34",
    diff_shock   = FALSE,
    panel_model  = "within",
    panel_effect = "individual",
    c_exog_data  = "Decade",
    robust_cov   = "vcovSCC",
    confint      = 1.96,
    hor          = HOR + 1
  )
  tibble(
    horizon  = 0:HOR,
    irf_mean = as.numeric(res$irf_panel_mean[1, 1:(HOR+1)]),
    irf_up   = as.numeric(res$irf_panel_up[1,   1:(HOR+1)]),
    irf_down = as.numeric(res$irf_panel_low[1,  1:(HOR+1)]),
    spec     = label
  )
}

irf_fixest_orig  <- run_fixest(fishprice_raw,            "fixest – original (calendar)")
irf_plm_orig     <- run_plm(fishprice_raw,               "plm – original")
irf_fixest_pos   <- run_fixest_positional(fishprice_raw, "fixest – positional lead (full sample)")
irf_fixest_fixed <- run_fixest(fishprice_fixed,          "fixest – drop ≥10yr gaps + balance")
irf_plm_fixed    <- run_plm(fishprice_fixed,             "plm – drop ≥10yr gaps + balance")
```

## Difference tables

```{r tables}
make_diff_table <- function(irf_a, irf_b, label_a, label_b) {
  inner_join(
    irf_a %>% dplyr::select(horizon, irf_mean) %>% rename(a = irf_mean),
    irf_b %>% dplyr::select(horizon, irf_mean) %>% rename(b = irf_mean),
    by = "horizon"
  ) %>%
    mutate(diff = round(a - b, 6),
           !!label_a := round(a, 6),
           !!label_b := round(b, 6)) %>%
    dplyr::select(horizon, !!label_a, !!label_b, diff)
}

cat("── Original: fixest calendar vs plm positional (should differ) ──\n")
print(make_diff_table(irf_fixest_orig,  irf_plm_orig,  "fixest_cal", "plm"))

cat("\n── fixest positional vs plm (full sample, should match) ──\n")
print(make_diff_table(irf_fixest_pos,   irf_plm_orig,  "fixest_pos", "plm"))

cat("\n── Robustness: drop ≥10yr gaps + balance ──\n")
print(make_diff_table(irf_fixest_fixed, irf_plm_fixed, "fixest_fix", "plm_fix"))
```

## Plot

```{r plot, fig.width=7, fig.height=4}
all_irf <- bind_rows(
  irf_fixest_orig, irf_plm_orig, irf_fixest_pos,
  irf_fixest_fixed, irf_plm_fixed
) %>%
  mutate(
    estimator = case_when(
      grepl("positional", spec) ~ "fixest positional",
      grepl("fixest",     spec) ~ "fixest calendar",
      TRUE                      ~ "plm"
    ),
    dataset = case_when(
      grepl("original|positional", spec) ~ "Full sample",
      TRUE                               ~ "Drop ≥10yr gaps + balance"
    ),
    dataset = factor(dataset, levels = c("Full sample", "Drop ≥10yr gaps + balance"))
  )

ggplot(all_irf, aes(horizon, irf_mean, colour = estimator, linetype = estimator)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up, fill = estimator),
              alpha = 0.10, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3, linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  facet_wrap(~dataset, nrow = 1) +
  scale_x_continuous(breaks = seq(0, HOR, 2)) +
  scale_colour_manual(values = c("fixest calendar" = "#333333",
                                  "fixest positional" = "#0072B2",
                                  "plm" = "#D55E00")) +
  scale_fill_manual(values   = c("fixest calendar" = "#333333",
                                  "fixest positional" = "#0072B2",
                                  "plm" = "#D55E00")) +
  scale_linetype_manual(values = c("fixest calendar" = "solid",
                                    "fixest positional" = "dashed",
                                    "plm" = "dotdash")) +
  labs(
    x = "Horizon (years)",
    y = "Effect on log fish prices",
    title = "fixest vs plm: before and after gap fix"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_blank(),
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  ) +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.2, alpha = 1)))
```

```{r}

```



---

## Optimal panel: maximal spatial & temporal coverage with minimal gaps

The goal is to select the best panel — one series per location — trading off
temporal span, spatial breadth, and gap severity.

```{r optimal-panel}
library(sf)

# ── Per-series coverage statistics ────────────────────────────────────────────
series_stats <- fishprice_raw %>%
  arrange(LocationSpecies, Year) %>%
  group_by(LocationSpecies, Location) %>%
  summarise(
    lat      = first(lat),
    lon      = first(lon),
    n_obs    = n(),
    min_yr   = min(Year),
    max_yr   = max(Year),
    span     = max(Year) - min(Year) + 1,
    n_gaps   = sum(diff(Year) > 1),
    max_gap  = if_else(n() > 1, max(diff(Year)) - 1L, 0L),
    gap_yrs  = sum(pmax(diff(Year) - 1, 0)),
    density  = n_obs / span,
    .groups  = "drop"
  )

# ── Score each series ──────────────────────────────────────────────────────────
# Objectives (all on 0-1 scale after normalisation):
#   + temporal span (longer is better)
#   + observation density (fewer gaps relative to span)
#   - max_gap (larger worst gaps penalised more heavily)
#   + n_obs (raw count as tiebreaker)

series_stats <- series_stats %>%
  mutate(
    score_span    = (span    - min(span))    / (max(span)    - min(span)),
    score_density = (density - min(density)) / (max(density) - min(density)),
    score_maxgap  = 1 - (max_gap - min(max_gap)) / (max(max_gap) - min(max_gap)),
    score_nobs    = (n_obs  - min(n_obs))    / (max(n_obs)    - min(n_obs)),
    # Weighted composite: penalise max_gap most heavily
    score = 0.30 * score_span +
            0.30 * score_density +
            0.30 * score_maxgap +
            0.10 * score_nobs
  ) %>%
  arrange(Location, desc(score))

cat("All series ranked within location:\n")
print(series_stats %>%
  dplyr::select(LocationSpecies, Location, span, n_obs, max_gap, gap_yrs, density, score) %>%
  mutate(across(c(density, score), ~ round(.x, 3))))

# ── Pick best series per location ─────────────────────────────────────────────
best_per_location <- series_stats %>%
  group_by(Location) %>%
  slice_max(score, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("\nOptimal panel (one series per location):\n")
print(best_per_location %>%
  dplyr::select(LocationSpecies, Location, lat, lon, span, n_obs, max_gap, gap_yrs, density, score) %>%
  mutate(across(c(density, score), ~ round(.x, 3))) %>%
  arrange(lat))
```

```{r optimal-map, fig.width=6, fig.height=4}
# ── Map of selected series ─────────────────────────────────────────────────────
world <- map_data("world") %>%
  filter(lat > 30, lat < 70, long > -15, long < 30)

ggplot() +
  geom_polygon(data = world, aes(long, lat, group = group),
               fill = "grey92", colour = "grey70", linewidth = 0.2) +
  geom_point(data = best_per_location,
             aes(lon, lat, size = span, colour = max_gap),
             shape = 21, stroke = 0.6, fill = "white") +
  geom_point(data = best_per_location,
             aes(lon, lat, size = span, colour = max_gap)) +
  ggrepel::geom_label_repel(data = best_per_location,
                             aes(lon, lat, label = Location),
                             size = 3, label.padding = unit(0.15, "lines"),
                             box.padding = 0.4, max.overlaps = 20) +
  scale_colour_viridis_c(option = "plasma", direction = -1,
                         name = "Max gap (yrs)") +
  scale_size_continuous(range = c(3, 9), name = "Temporal span (yrs)") +
  coord_fixed(xlim = c(-12, 26), ylim = c(35, 62)) +
  labs(x = NULL, y = NULL,
       title = "Optimal panel: best series per location",
       subtitle = "Size = temporal span, colour = max gap (lower = better)") +
  theme_classic(base_size = 11) +
  theme(legend.position = "right")
```

```{r optimal-timeline, fig.width=7, fig.height=3.5}
# ── Timeline of selected series ───────────────────────────────────────────────
timeline_df <- fishprice_raw %>%
  filter(LocationSpecies %in% best_per_location$LocationSpecies) %>%
  group_by(LocationSpecies, Location) %>%
  summarise(min_yr = min(Year), max_yr = max(Year), .groups = "drop") %>%
  left_join(best_per_location %>% dplyr::select(LocationSpecies, max_gap, score), by = "LocationSpecies") %>%
  arrange(lat) %>%
  left_join(best_per_location %>% dplyr::select(LocationSpecies, lat), by = "LocationSpecies") %>%
  mutate(Location = factor(Location, levels = Location))

ggplot(timeline_df, aes(y = Location, colour = max_gap)) +
  geom_segment(aes(x = min_yr, xend = max_yr, yend = Location), linewidth = 3) +
  # mark actual observed years as ticks
  geom_point(data = fishprice_raw %>%
               filter(LocationSpecies %in% best_per_location$LocationSpecies) %>%
               left_join(best_per_location %>% dplyr::select(LocationSpecies, max_gap), by = "LocationSpecies"),
             aes(x = Year, y = Location),
             shape = "|", size = 1.5, alpha = 0.4) +
  scale_colour_viridis_c(option = "plasma", direction = -1, name = "Max gap (yrs)") +
  scale_x_continuous(breaks = seq(1500, 1800, 50)) +
  labs(x = "Year", y = NULL,
       title = "Temporal coverage of optimal panel") +
  theme_classic(base_size = 11) +
  theme(legend.position = "right")
```

```{r optimal-irf, cache=TRUE, fig.width=6, fig.height=3.5}
# ── IRF on optimal panel (calendar fixest, gold standard) ─────────────────────
fishprice_optimal <- fishprice_raw %>%
  filter(LocationSpecies %in% best_per_location$LocationSpecies)

irf_optimal <- lp_panel(
  data         = fishprice_optimal,
  outcome      = "logprice",
  main_var     = "nino34",
  controls     = NULL,
  horizon      = HOR,
  fe           = "LocationSpecies + Decade",
  panel_id     = c("LocationSpecies", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(spec = "Optimal panel (calendar fixest)")

# Compare to full-sample original
ggplot(bind_rows(irf_fixest_orig, irf_optimal),
       aes(horizon, irf_mean, colour = spec, fill = spec)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.12, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3, linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(breaks = seq(0, HOR, 2)) +
  scale_colour_manual(values = c("fixest – original (calendar)" = "#333333",
                                  "Optimal panel (calendar fixest)" = "#0072B2")) +
  scale_fill_manual(values   = c("fixest – original (calendar)" = "#333333",
                                  "Optimal panel (calendar fixest)" = "#0072B2")) +
  labs(x = "Horizon (years)", y = "Effect on log fish prices",
       title = "Full sample vs optimal panel IRF") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom", legend.title = element_blank(),
        strip.background = element_blank()) +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.2, alpha = 1)))
```

```{r loo-full, cache=TRUE, fig.width=7, fig.height=4}
# ── Leave-one-location-out: full sample ───────────────────────────────────────
locations_full <- unique(fishprice_raw$Location)

loo_full <- map_dfr(locations_full, function(loc) {
  lp_panel(
    data         = fishprice_raw %>% filter(Location != loc),
    outcome      = "logprice",
    main_var     = "nino34",
    controls     = NULL,
    horizon      = HOR,
    fe           = "LocationSpecies + Decade",
    panel_id     = c("LocationSpecies", "Year"),
    vcov_formula = DK ~ Year
  ) %>% mutate(left_out = loc)
})

ggplot(loo_full, aes(horizon, irf_mean, colour = left_out, fill = left_out, group = left_out)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.07, colour = NA) +
  geom_line(linewidth = 0.6) +
  # Baseline: full sample
  geom_line(data = irf_fixest_orig, aes(horizon, irf_mean),
            colour = "black", linewidth = 1.1, linetype = "dashed", inherit.aes = FALSE) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0, HOR, 2)) +
  scale_colour_viridis_d(option = "turbo", name = "Left out") +
  scale_fill_viridis_d(option = "turbo", name = "Left out") +
  labs(x = "Horizon (years)", y = "Effect on log fish prices",
       title = "Leave-one-location-out IRF – full sample",
       subtitle = "Dashed black = full sample baseline") +
  theme_classic(base_size = 12) +
  theme(legend.position = "right", strip.background = element_blank())
```

```{r loo-optimal, cache=TRUE, fig.width=7, fig.height=4}
# ── Leave-one-location-out: optimal panel ─────────────────────────────────────
locations_optimal <- unique(fishprice_optimal$Location)

loo_optimal <- map_dfr(locations_optimal, function(loc) {
  lp_panel(
    data         = fishprice_optimal %>% filter(Location != loc),
    outcome      = "logprice",
    main_var     = "nino34",
    controls     = NULL,
    horizon      = HOR,
    fe           = "LocationSpecies + Decade",
    panel_id     = c("LocationSpecies", "Year"),
    vcov_formula = DK ~ Year
  ) %>% mutate(left_out = loc)
})

ggplot(loo_optimal, aes(horizon, irf_mean, colour = left_out, fill = left_out, group = left_out)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.07, colour = NA) +
  geom_line(linewidth = 0.6) +
  geom_line(data = irf_optimal, aes(horizon, irf_mean),
            colour = "black", linewidth = 1.1, linetype = "dashed", inherit.aes = FALSE) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0, HOR, 2)) +
  scale_colour_viridis_d(option = "turbo", name = "Left out") +
  scale_fill_viridis_d(option = "turbo", name = "Left out") +
  labs(x = "Horizon (years)", y = "Effect on log fish prices",
       title = "Leave-one-location-out IRF – optimal panel",
       subtitle = "Dashed black = optimal panel baseline") +
  theme_classic(base_size = 12) +
  theme(legend.position = "right", strip.background = element_blank())
```

---

## Species-interacted IRFs

```{r species-inter, cache=TRUE, fig.width=7, fig.height=4}
# ── Species-interacted IRF: full sample & no-max-gap-10 ───────────────────────
# fishprice_nogap: drop series with max_gap > 10 (slightly looser than the
# gap-fix threshold, so that Munich Cod+Herring — both with max_gap <= 10 —
# are retained, ensuring at least one dual-species location for identification)
nogap_series <- gap_summary %>% filter(max_gap <= 15) %>% pull(LocationSpecies)
fishprice_nogap <- fishprice_raw %>%
  filter(LocationSpecies %in% nogap_series)

run_inter <- function(df, label) {
  lp_panel_inter(
    data         = df,
    outcome      = "logprice",
    main_var     = "nino34",
    interact_var = "Species",
    controls     = NULL,
    horizon      = HOR,
    fe           = "LocationSpecies + Decade",
    panel_id     = c("LocationSpecies", "Year"),
    vcov_formula = DK ~ Year
  ) %>% mutate(dataset = label)
}

irf_inter_full  <- run_inter(fishprice_raw,   "Full sample")
irf_inter_nogap <- run_inter(fishprice_nogap, "No max-gap > 10 yrs")

irf_inter <- bind_rows(irf_inter_full, irf_inter_nogap) %>%
  mutate(
    dataset  = factor(dataset, levels = c("Full sample", "No max-gap > 10 yrs")),
    category = factor(category, levels = c("Cod", "Herring"))
  )

ggplot(irf_inter, aes(horizon, irf_mean, colour = category, fill = category)) +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.13, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3, linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  facet_wrap(~dataset, nrow = 1) +
  scale_x_continuous(breaks = seq(0, HOR, 2)) +
  scale_colour_manual(values = c(Cod = "#0072B2", Herring = "#D55E00")) +
  scale_fill_manual(values   = c(Cod = "#0072B2", Herring = "#D55E00")) +
  labs(
    x      = "Horizon (years)",
    y      = "Effect on log fish prices",
    colour = "Species", fill = "Species",
    title  = "Species-interacted IRF: ENSO × Cod vs Herring"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position  = "bottom",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  ) +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.2, alpha = 1)))
```

```{r species-inter-loo, cache=TRUE, fig.width=9, fig.height=8}
# ── LOO × species interaction ─────────────────────────────────────────────────
# For each dataset: leave one location out, re-run interacted IRF,
# collect all curves — shows which location drives the Cod vs Herring split

run_loo_inter <- function(df, locations, label) {
  map_dfr(locations, function(loc) {
    tryCatch(
      lp_panel_inter(
        data         = df %>% filter(Location != loc),
        outcome      = "logprice",
        main_var     = "nino34",
        interact_var = "Species",
        controls     = NULL,
        horizon      = HOR,
        fe           = "LocationSpecies + Decade",
        panel_id     = c("LocationSpecies", "Year"),
        vcov_formula = DK ~ Year
      ) %>% mutate(left_out = loc, dataset = label),
      error = function(e) NULL
    )
  })
}

loo_inter_full  <- run_loo_inter(fishprice_raw,   unique(fishprice_raw$Location),   "Full sample")
loo_inter_nogap <- run_loo_inter(fishprice_nogap, unique(fishprice_nogap$Location), "No max-gap > 10 yrs")

loo_inter <- bind_rows(loo_inter_full, loo_inter_nogap) %>%
  mutate(
    dataset  = factor(dataset, levels = c("Full sample", "No max-gap > 10 yrs")),
    category = factor(category, levels = c("Cod", "Herring"))
  )

ggplot(loo_inter, aes(horizon, irf_mean, colour = left_out, group = left_out)) +
  geom_line(linewidth = 0.5, alpha = 0.7) +
  # Baseline interacted IRF (all locations)
  geom_line(
    data = irf_inter,
    aes(horizon, irf_mean, group = category),
    colour = "black", linewidth = 1.0, linetype = "dashed", inherit.aes = FALSE
  ) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3, linetype = "dotted") +
  facet_grid(category ~ dataset) +
  scale_x_continuous(breaks = seq(0, HOR, 2)) +
  scale_colour_viridis_d(option = "turbo", name = "Left out") +
  labs(
    x      = "Horizon (years)",
    y      = "Effect on log fish prices",
    title  = "LOO × species interaction",
    subtitle = "Dashed black = baseline with all locations"
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position  = "right",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

```{r loo-compare, fig.width=8, fig.height=4}
# ── Side-by-side: which location drives the result? ───────────────────────────
# Summarise influence as max absolute deviation from baseline at any horizon
baseline_full    <- irf_fixest_orig %>% dplyr::select(horizon, base = irf_mean)
baseline_optimal <- irf_optimal     %>% dplyr::select(horizon, base = irf_mean)

influence_full <- loo_full %>%
  left_join(baseline_full, by = "horizon") %>%
  group_by(left_out) %>%
  summarise(max_dev = max(abs(irf_mean - base)), .groups = "drop") %>%
  mutate(panel = "Full sample")

influence_optimal <- loo_optimal %>%
  left_join(baseline_optimal, by = "horizon") %>%
  group_by(left_out) %>%
  summarise(max_dev = max(abs(irf_mean - base)), .groups = "drop") %>%
  mutate(panel = "Optimal panel")

influence <- bind_rows(influence_full, influence_optimal) %>%
  mutate(panel = factor(panel, levels = c("Full sample", "Optimal panel")))

ggplot(influence, aes(reorder(left_out, max_dev), max_dev, fill = panel)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("Full sample" = "#333333", "Optimal panel" = "#0072B2")) +
  labs(x = NULL, y = "Max |deviation from baseline| across horizons",
       title = "Location influence on IRF",
       fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```
