---
title: "Volcanic Forcing – ENSO Correlation and Price/Famine Analyses"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

Does volcanic forcing confound ENSO effects on grain prices, fish prices, and
famine onset? This notebook:

1. Loads and cleans the volcanic forcing dataset
2. Merges it with grain price, fish price, and famine onset data
3. Checks correlation between ENSO and volcanic forcing
4. Re-runs LP-IRF analyses controlling for volcanic forcing

## Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(fixest)
  library(readxl)
  library(purrr)
  library(broom)
  library(scales)
})

source("../emileRegs.R")
setFixest_notes(FALSE)

# ── Colour palette ────────────────────────────────────────────────────────────
COL_ENSO  <- "dodgerblue"       # ENSO series
COL_GVF   <- "tomato"           # GVF series
COL_GR    <- "#0F3460"          # deep blue
COL_ANT   <- "#E2AD1A"          # amber
COL_BASE  <- "dodgerblue"       # baseline IRF
COL_VOLC  <- "tomato"           # volcanic-controlled IRF

nature_theme <- function(base_size = 13) {
  theme_classic(base_size = base_size) +
    theme(
      axis.line         = element_line(linewidth = 0.5, colour = "black"),
      axis.ticks        = element_line(linewidth = 0.4, colour = "black"),
      axis.ticks.length = unit(3, "pt"),
      axis.text         = element_text(size = base_size - 1, colour = "black",
                                       face = "bold"),
      axis.title        = element_text(size = base_size, colour = "black",
                                       face = "bold"),
      legend.position   = "bottom",
      legend.key.size   = unit(10, "pt"),
      legend.text       = element_text(size = base_size - 1, face = "bold"),
      legend.title      = element_blank(),
      legend.margin     = margin(0, 0, 0, 0),
      legend.spacing.x  = unit(5, "pt"),
      strip.background  = element_blank(),
      strip.text        = element_text(face = "bold", size = base_size),
      plot.tag          = element_text(face = "bold", size = base_size + 4,
                                       colour = "black"),
      plot.tag.position = "topleft",
      plot.margin       = margin(6, 8, 6, 8, "pt"),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank()
    )
}
```

## 1. Load volcanic forcing data

```{r volcanic-load}
raw <- read_excel("../data/volcanic_data.xlsx",
                  sheet = "1 - Volcanic_Forcing",
                  col_names = FALSE)

# Row 1 = header level 1, row 2 = header level 2, rows 3+ = data
# Columns (1-indexed in R):
#   1  Year
#   2  Source (1=tropical, 2=NH, 3=SH)
#   3  Greenland sulfate [kg km-2]
#   4  Greenland uncertainty
#   5  Antarctica sulfate [kg km-2]
#   6  Antarctica uncertainty
#   7  Global Volcanic Forcing GVF [W m-2]
#   8  GVF uncertainty

volc_raw <- raw[-c(1, 2), 1:8]
colnames(volc_raw) <- c(
  "Year", "Source",
  "greenland_sulfate", "greenland_unc",
  "antarctica_sulfate", "antarctica_unc",
  "GVF", "GVF_unc"
)

volc_raw <- volc_raw %>%
  mutate(across(everything(), as.numeric)) %>%
  filter(!is.na(Year), Year >= 1500, Year <= 1900) %>%
  arrange(Year)

cat("Volcanic events in sample (1500-1900):", nrow(volc_raw), "\n")
print(volc_raw)
```

## 2. Build annual volcanic forcing series

```{r volcanic-annual}
year_grid <- tibble(Year = 1500:1900)

volc <- year_grid %>%
  left_join(volc_raw, by = "Year") %>%
  mutate(
    GVF                = replace_na(GVF, 0),
    greenland_sulfate  = replace_na(greenland_sulfate, 0),
    antarctica_sulfate = replace_na(antarctica_sulfate, 0),
    eruption           = as.integer(GVF != 0)
  ) %>%
  dplyr::select(Year, GVF, greenland_sulfate, antarctica_sulfate, eruption) %>%
  mutate(
    GVF_l1 = lag(GVF, 1),
    GVF_l2 = lag(GVF, 2),
    GVF_l3 = lag(GVF, 3)
  )

summary(volc[, c("GVF", "greenland_sulfate", "antarctica_sulfate")])
```

## 3. Load main datasets and merge

```{r data-load}
price <- read.csv("../processed data/price_2023_enso.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  drop_na() %>% distinct() %>% arrange(Location, Year)

fishprice <- read.csv("../processed data/fishprice_enso.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  arrange(LocationSpecies, Year)

famines <- read.csv("../processed data/famine_region_data.csv") %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  arrange(Region, Year)

price     <- left_join(price,     volc, by = "Year")
fishprice <- left_join(fishprice, volc, by = "Year")
famines   <- left_join(famines,   volc, by = "Year")

cat("Grain price obs after merge:", nrow(price), "\n")
cat("Fish price obs after merge: ", nrow(fishprice), "\n")
cat("Famine obs after merge:     ", nrow(famines), "\n")
```

## 4. ENSO × Volcanic forcing correlation

```{r enso-volc-corr}
ts <- price %>%
  group_by(Year) %>%
  summarise(
    ENSO               = mean(nino34,             na.rm = TRUE),
    GVF                = mean(GVF,                na.rm = TRUE),
    greenland_sulfate  = mean(greenland_sulfate,  na.rm = TRUE),
    antarctica_sulfate = mean(antarctica_sulfate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(c(ENSO, GVF, greenland_sulfate, antarctica_sulfate),
                ~ as.numeric(scale(.x))))

fmt_p <- function(p) if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)

cor_gvf  <- cor.test(ts$ENSO, ts$GVF)
cor_gr   <- cor.test(ts$ENSO, ts$greenland_sulfate)
cor_ant  <- cor.test(ts$ENSO, ts$antarctica_sulfate)

cat(sprintf("ENSO vs GVF:                r = %.3f, %s\n",
            cor_gvf$estimate,  fmt_p(cor_gvf$p.value)))
cat(sprintf("ENSO vs Greenland sulfate:  r = %.3f, %s\n",
            cor_gr$estimate,   fmt_p(cor_gr$p.value)))
cat(sprintf("ENSO vs Antarctica sulfate: r = %.3f, %s\n",
            cor_ant$estimate,  fmt_p(cor_ant$p.value)))
```

## 5. Grain price IRFs

```{r grain-irf, cache=TRUE}
run_lp <- function(df, controls, fe, panel_id) {
  lp_panel(
    data         = df,
    outcome      = "logprice",
    main_var     = "nino34",
    controls     = controls,
    horizon      = 10,
    fe           = fe,
    panel_id     = panel_id,
    vcov_formula = DK ~ Year
  )
}

irf_grain_base <- run_lp(
  price,
  controls = "i(Decade)",
  fe       = "Location",
  panel_id = c("Location", "Year")
) %>% mutate(spec = "Baseline")

irf_grain_volc <- run_lp(
  price %>% drop_na(GVF),
  controls = "i(Decade) + GVF + greenland_sulfate + antarctica_sulfate ",
  fe       = "Location",
  panel_id = c("Location", "Year")
) %>% mutate(spec = "Controlling for V.F.")

irf_grain <- bind_rows(irf_grain_base, irf_grain_volc) %>%
  mutate(spec = factor(spec, levels = c("Baseline", "Controlling for V.F.")))
```



## 7. Famine onset – region interaction model

```{r famine-model}
m_base <- feols(
  Famine_start ~ nino34:i(Region) | Region + Decade,
  data     = famines,
  panel.id = c("Region", "Year"),
  vcov     = DK ~ Year
)

m_volc <- feols(
  Famine_start ~ nino34:i(Region) + GVF + greenland_sulfate + antarctica_sulfate | Region + Decade,
  data     = famines %>% drop_na(GVF),
  panel.id = c("Region", "Year"),
  vcov     = DK ~ Year
)

etable(m_base, m_volc,
       dict    = c("nino34" = "NINO3.4", "GVF" = "GVF T",
                   "GVF_l1" = "GVF T-1", "GVF_l2" = "GVF T-2",
                   "GVF_l3" = "GVF T-3"),
       fitstat = c("n", "r2"))
```

```{r famine-coefs}
# Extract region coefficients with 90 and 95 % CIs
extract_region_coefs <- function(mod, spec_label) {
  b   <- coef(mod)
  V   <- vcov(mod)
  nms <- names(b)
  idx <- grep("nino34", nms)

  tibble(
    term     = nms[idx],
    estimate = b[idx],
    se       = sqrt(diag(V)[idx]),
    spec     = spec_label
  ) %>%
    mutate(
      lo95 = estimate - 1.960 * se,
      hi95 = estimate + 1.960 * se,
      lo90 = estimate - 1.645 * se,
      hi90 = estimate + 1.645 * se,
      Region = term %>%
        gsub("nino34:Region::", "", .) %>%
        gsub("nino34:i\\(Region\\)::", "", .) %>%
        trimws()
    )
}

coef_base <- extract_region_coefs(m_base, "Baseline")
coef_volc <- extract_region_coefs(m_volc, "Controlling for V.F.")
coef_df   <- bind_rows(coef_base, coef_volc) %>%
  mutate(spec = factor(spec, levels = c("Baseline", "Controlling for V.F.")))

print(coef_df[, c("Region", "spec", "estimate", "lo95", "hi95")])
```

## 8. Figure panels

### Panel a – Time series with correlation annotations

```{r panel-a, fig.width=8, fig.height=4.5}
# Build annotation labels
make_cor_label <- function(cor_obj, name) {
  sprintf("%s:  r = %.2f  (%s)", name,
          cor_obj$estimate, fmt_p(cor_obj$p.value))
}

ann_labels <- paste(
  make_cor_label(cor_gvf,  "GVF"),
  make_cor_label(cor_gr,   "Greenland"),
  make_cor_label(cor_ant,  "Antarctica"),
  sep = "\n"
)

ts_long <- ts %>%
  pivot_longer(-Year, names_to = "Index", values_to = "value") %>%
  mutate(Index = factor(Index,
    levels = c("ENSO", "GVF", "greenland_sulfate", "antarctica_sulfate"),
    labels = c("ENSO (NINO3.4)", "Global V.F. (GVF)",
               "Greenland sulfate", "Antarctica sulfate")))

panel_a <- ggplot(ts_long, aes(Year, value, colour = Index, linewidth = Index)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3, linetype = "dashed") +
  geom_line(alpha = 0.88) +
  annotate("label",
           x = 1895, y = max(ts_long$value, na.rm = TRUE) * 0.97,
           label     = ann_labels,
           hjust     = 1, vjust = 1,
           size      = 4, fontface = "bold",
           fill      = "white", colour = "grey20",
           label.size = 0.3, label.padding = unit(4, "pt")) +
  scale_colour_manual(
    values = c("ENSO (NINO3.4)"     = COL_ENSO,
               "Global V.F. (GVF)"  = COL_GVF,
               "Greenland sulfate"  = COL_GR,
               "Antarctica sulfate" = COL_ANT)
  ) +
  scale_linewidth_manual(
    values = c("ENSO (NINO3.4)"     = 0.9,
               "Global V.F. (GVF)"  = 0.7,
               "Greenland sulfate"  = 0.7,
               "Antarctica sulfate" = 0.7)
  ) +
  guides(linewidth = "none",
         colour = guide_legend(override.aes = list(linewidth = 1.4))) +
  scale_x_continuous(breaks = seq(1500, 1900, 50)) +
  labs(x = "Year", y = "Standardised value",
       title = "ENSO and volcanic forcing") +
  nature_theme()

panel_a
```

### Panel b – Grain price IRF overlay

```{r panel-b, fig.width=5, fig.height=4.5}
panel_b <- ggplot(irf_grain,
                  aes(horizon, irf_mean,
                      colour   = spec, fill = spec, linetype = spec)) +
  geom_hline(yintercept = 0, colour = "grey65", linewidth = 1.2, linetype = "dashed") +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = seq(0, 10, 2), expand = c(0.02, 0)) +
  scale_colour_manual(values   = c(COL_BASE, COL_VOLC)) +
  scale_fill_manual(values     = c(COL_BASE, COL_VOLC)) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  guides(colour   = guide_legend(override.aes = list(linewidth = 1.4)),
         fill     = guide_legend(override.aes = list(alpha = 0.2)),
         linetype = guide_legend(override.aes = list(linewidth = 1.4))) +
  labs(x = "Horizon (years)", y = "% Grain price response",
       title = "ENSO effect on grain prices") +
  nature_theme()

panel_b
```

### Panel c – Famine onset by region (Central Europe highlighted)

```{r panel-c, fig.width=5.5, fig.height=4.5}
# Dodge position for two specs
pd <- position_dodge(width = 0.55)

panel_c <- ggplot(coef_df,
                  aes(x = Region, y = estimate,
                      colour = spec, shape = spec)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.35,
             linetype = "dashed") +
  # 95 % CI (thin whisker)
  geom_errorbar(aes(ymin = lo95, ymax = hi95),
                width = 0.0, linewidth = 0.55,
                position = pd) +
  # 90 % CI (thick whisker)
  geom_errorbar(aes(ymin = lo90, ymax = hi90),
                width = 0.0, linewidth = 1.5,
                position = pd) +
  geom_point(size = 2.2, position = pd) +
  # Highlight Central Europe with a shaded band
  annotate("rect",
           xmin = 0.5, xmax = 1.5,
           ymin = -Inf, ymax = Inf,
           fill = "#E94560", alpha = 0.07) +
  scale_colour_manual(values = c(COL_BASE, COL_VOLC)) +
  scale_shape_manual(values  = c(16, 17)) +
  guides(colour = guide_legend(override.aes = list(size = 2.5)),
         shape  = guide_legend(override.aes = list(size = 2.5))) +
  labs(x = NULL, y = "ENSO effect on famine onset prob.",
       title = "Famine onset by region") +
  nature_theme() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 10))

panel_c
```

## 9. Combined figure 1 – ENSO robustness (a / b | c)

```{r combined, fig.width=12, fig.height=9}
fig <- panel_a / (panel_b | panel_c) +
  plot_layout(heights = c(1, 1.4)) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag   = element_text(face = "bold", size = 20, colour = "black"),
        plot.title = element_text(face = "bold", size = 15))

fig
```

## 10. Volcanic forcing as the shock – IRF on grain prices

```{r volc-irf, cache=TRUE}
# IRF of GVF on grain prices controlling for ENSO
irf_gvf_grain <- lp_panel(
  data         = price,
  outcome      = "logprice",
  main_var     = "GVF",
  controls     = "i(Decade) + nino34",
  horizon      = 10,
  fe           = "Location",
  panel_id     = c("Location", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(forcing = "Global V.F. (GVF)")

# Also run for Greenland and Antarctica separately
irf_gvf_gr <- lp_panel(
  data         = price,
  outcome      = "logprice",
  main_var     = "greenland_sulfate",
  controls     = "i(Decade) + nino34",
  horizon      = 10,
  fe           = "Location",
  panel_id     = c("Location", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(forcing = "Greenland sulfate")

irf_gvf_ant <- lp_panel(
  data         = price,
  outcome      = "logprice",
  main_var     = "antarctica_sulfate",
  controls     = "i(Decade) + nino34",
  horizon      = 10,
  fe           = "Location",
  panel_id     = c("Location", "Year"),
  vcov_formula = DK ~ Year
) %>% mutate(forcing = "Antarctica sulfate")

irf_volc_all <- bind_rows(irf_gvf_grain, irf_gvf_gr, irf_gvf_ant) %>%
  mutate(forcing = factor(forcing,
    levels = c("Global V.F. (GVF)", "Greenland sulfate", "Antarctica sulfate")))
```

```{r volc-irf-plot, fig.width=7, fig.height=4.5}
COL_VOLC3 <- c("Global V.F. (GVF)"  = "tomato",
               "Greenland sulfate"  = COL_GR,
               "Antarctica sulfate" = COL_ANT)

panel_d <- ggplot(irf_volc_all,
                  aes(horizon, irf_mean,
                      colour = forcing, fill = forcing)) +
  geom_hline(yintercept = 0, colour = "grey65", linewidth = 0.3, linetype = "dashed") +
  geom_ribbon(aes(ymin = irf_down, ymax = irf_up), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1.0) +
  scale_x_continuous(breaks = seq(0, 10, 2), expand = c(0.02, 0)) +
  scale_colour_manual(values = COL_VOLC3) +
  scale_fill_manual(values   = COL_VOLC3) +
  scale_y_continuous(labels  = percent_format(accuracy = 0.1)) +
  guides(colour = guide_legend(override.aes = list(linewidth = 1.4)),
         fill   = guide_legend(override.aes = list(alpha = 0.2))) +
  labs(x = "Horizon (years)", y = "% Grain price response",
       title = "V.F. effect on grain prices (controlling for ENSO)") +
  nature_theme()

panel_d
```

## 11. Volcanic forcing – famine onset by region (controlling for ENSO)



```{r volc-famine-coefs}
m_volc_region <- feols(
  Famine_start ~ GVF:i(Region) + nino34 | Region + Decade,
  data     = famines,
  panel.id = c("Region", "Year"),
  vcov     = DK ~ Year
)

# Extract GVF × Region coefficients
b_vr   <- coef(m_volc_region)
V_vr   <- vcov(m_volc_region)
idx_vr <- grep("GVF", names(b_vr))

coef_vr <- tibble(
  term     = names(b_vr)[idx_vr],
  estimate = b_vr[idx_vr],
  se       = sqrt(diag(V_vr)[idx_vr])
) %>%
  mutate(
    lo95 = estimate - 1.960 * se,
    hi95 = estimate + 1.960 * se,
    lo90 = estimate - 1.645 * se,
    hi90 = estimate + 1.645 * se,
    Region = term %>%
      gsub("GVF:i\\(Region\\)::", "", .) %>%
      gsub("GVF:Region::",        "", .) %>%
      trimws()
  )

print(coef_vr[, c("Region", "estimate", "lo95", "hi95")])
```

```{r volc-famine-plot, fig.width=5.5, fig.height=4.5}
panel_e <- ggplot(coef_vr, aes(x = Region, y = estimate)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.35,
             linetype = "dashed") +
  geom_errorbar(aes(ymin = lo95, ymax = hi95),
                width = 0.0, linewidth = 0.55, colour = "tomato") +
  geom_errorbar(aes(ymin = lo90, ymax = hi90),
                width = 0.0, linewidth = 1.5, colour = "tomato") +
  geom_point(size = 2.4, colour = "tomato", shape = 17) +
  annotate("rect",
           xmin = 0.5, xmax = 1.5,
           ymin = -Inf, ymax = Inf,
           fill = "tomato", alpha = 0.07) +
  labs(x = NULL, y = "V.F. effect on famine onset prob.",
       title = "V.F. effect by region (controlling for ENSO)") +
  nature_theme() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 10))

panel_e
```

## 12. Combined figure 2 – Volcanic forcing as shock (d | e)

```{r combined2, fig.width=12, fig.height=5}
fig2 <- (panel_d | panel_e) +
  plot_layout(widths = c(1.4, 1.2)) +
  plot_annotation(tag_levels = list(c("d", "e"))) &
  theme(plot.tag   = element_text(face = "bold", size = 20, colour = "black"),
        plot.title = element_text(face = "bold", size = 13))

fig2
```

## Save

```{r save}
out_dir <- "../analysis/output/figures/extended data"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(
  file.path(out_dir, "figED_volcanic.pdf"),
  plot   = fig,
  width  = 12,
  height = 9,
  units  = "in",
  device = cairo_pdf
)

ggsave(
  file.path(out_dir, "figED_volcanic2.pdf"),
  plot   = fig2,
  width  = 12,
  height = 5,
  units  = "in",
  device = cairo_pdf
)

message("Saved → figED_volcanic.pdf and figED_volcanic2.pdf")
```
